---
title: "DataSHIELD Workshop: Data Science Without Borders"
freeze: true
format: 
  html:
    code-copy: false
execute:
  eval: true
  echo: true
  warning: false
  error: true
---

<a href="#" onclick="event.preventDefault(); fetch('https://raw.githubusercontent.com/isglobal-brge/workshop_DSWB/main/DataSHIELD_workshop_DSWB_resources.R').then(response => response.blob()).then(blob => { const url = window.URL.createObjectURL(blob); const a = document.createElement('a'); a.style.display = 'none'; a.href = url; a.download = 'DataSHIELD_workshop_DSWB_resources.R'; document.body.appendChild(a); a.click(); window.URL.revokeObjectURL(url); }); return false;" class="btn btn-primary btn-sm">
<i class="bi bi-download"></i> Download this R file
</a>

# Analysis using the `resources`

For more examples and detailed information about working with resources in DataSHIELD, please refer to the [Resources Workshop Repository](https://github.com/obiba/resources-workshop).

Now, let us illustrate a similar analysis of the previous example using CNSIM datasets but having the data as a resource. Now the resources are available in a project called `RSRC` (see https://opal-demo.obiba.org/#/project/RSRC/resources). Now, we write all the require code in a single chunk:

```{r cnsim_multiple}
library(DSOpal)
library(dsBaseClient)

# prepare login data and resources to assign
builder <- DSI::newDSLoginBuilder()
builder$append(server = "study1", url = "https://opal-demo.obiba.org", 
               user = "dsuser", password = "P@ssw0rd", 
               resource = "RSRC.CNSIM1", profile = "default")
#builder$append(server = "study2", url = "https://opal.isglobal.org/repo",
#               user = "invited",  password = "12345678Aa@",, 
#               resource = "CNSIM.CNSIM2", profile = "rock-inma")

logindata <- builder$build()

# login and assign resources
conns <- datashield.login(logins = logindata, assign = TRUE, symbol = "res")

# assigned objects are of class ResourceClient (and others)
ds.class("res")

# coerce ResourceClient objects to data.frames
# (DataSHIELD config allows as.resource.data.frame() assignment function for the purpose of the demo)
datashield.assign.expr(conns, symbol = "D", 
                       expr = quote(as.resource.data.frame(res, strict = TRUE)))
ds.class("D")
ds.colnames("D")

# do usual dsBase analysis
ds.summary('D$LAB_HDL')

# vector types are not necessarily the same depending on the data reader that was used
ds.class('D$GENDER')
ds.asFactor('D$GENDER', 'GENDER')
ds.summary('GENDER')

mod <- ds.glm("DIS_DIAB ~ LAB_TRIG + GENDER", data = "D" , family="binomial")
mod$coeff


datashield.logout(conns)
```

# Survival analysis

Survival analysis can be performed in DataSHIELD using `dsSurvival` package 
(paper [https://bmcresnotes.biomedcentral.com/articles/10.1186/s13104-022-06085-1](https://bmcresnotes.biomedcentral.com/articles/10.1186/s13104-022-06085-1))

Install the package which allows you to perform survival analysis in DataSHIELD

```{r install_dsSurvivalClient, eval=FALSE}
devtools::install_github("neelsoumya/dsSurvivalClient")
```

Login to the Opal server which is having CORDELIA data. 
IMPORTANT: In order to have an Opal instance with dsSurvival package available, 
use the profile "lemon-donkey"

Now, read how to perform surival analysis using DataSHIELD in this vignette

[https://neelsoumya.github.io/dsSurvivalbookdown/](https://neelsoumya.github.io/dsSurvivalbookdown/)
  
TASK: Perform survival analysis of CORDELIA dataset where the time until death 
is in the variable called "todeath" and censoring variable is in the variable 
"death" to investigate whether smoking (variable "fumador") is associated or not
with mortality after adjusting for age (variable "edad")
NOTE: follow all steps in the vignette.

# Omic data analysis

Please refer to the [Omic data analysis resource](https://isglobal-brge.github.io/resource_bookdown/Omic.html).

# Exercise: CORDELIA Project Analysis

We have a database from the CORDELIA Project (https://cordeliaproject.net/), which collects clinical, demographic, and laboratory information about participants in order to study cardiovascular and metabolic risk factors. The dataset contains the following variables:

- **cohorte** – Cohort (study group the subject belongs to)
- **codigo** – Code / ID (unique identifier for each subject)
- **fec_inc** – Date of inclusion (enrollment date in the study)
- **edad** – Age
- **sexo** – Sex (male/female)
- **est_civ** – Marital status
- **niv_cult** – Educational level
- **act_fis** – Physical activity
- **fumador** – Smoker (yes/no)
- **HTA** – Hypertension (diagnosis)
- **HTA_TTO** – Hypertension treatment (yes/no)
- **hipercol** – Hypercholesterolemia (diagnosis)
- **col_tto** – Cholesterol treatment (yes/no)
- **diabetes** – Diabetes (diagnosis)
- **diab_tto** – Diabetes treatment (yes/no)
- **insulin** – Insulin treatment (yes/no)
- **diab_ado_ins** – Diabetes treated with oral antidiabetics and insulin
- **peso** – Weight (kg)
- **talla** – Height (cm or m)
- **cintura** – Waist circumference (cm)
- **IMC** – Body Mass Index (BMI)
- **FC** – Heart rate (beats per minute)
- **PAS_1** – Systolic blood pressure (1st measurement)
- **PAS_2** – Systolic blood pressure (2nd measurement)
- **PAD_1** – Diastolic blood pressure (1st measurement)
- **PAD_2** – Diastolic blood pressure (2nd measurement)
- **col_tot** – Total cholesterol
- **hdl** – HDL cholesterol
- **ldl** – LDL cholesterol
- **trig** – Triglycerides
- **creat** – Creatinine
- **filt_glomer** – Glomerular filtration rate (GFR)
- **glu** – Blood glucose
- **exitusSeg** – Death during follow-up (yes/no)
- **FechaExit_Seg** – Date of death during follow-up
- **filt_glomer_CKD** – Glomerular filtration rate classified for chronic kidney disease (CKD stages)
- **HTA_tot** – Hypertension (overall variable, combining diagnosis/treatment)
- **diabetes_tot** – Diabetes (overall variable, combining diagnosis/treatment)
- **todeath** – Time to death (follow-up time until death)
- **death** – Death (event indicator, yes/no)

The data is stored on an Opal server within a project named `CORDELIA`, under the resource `cordelia45`, with access credentials provided as follows:

- Server: "https://opal-demo.obiba.org"
- User: "dsuser"
- Password: "P@ssw0rd"

## Tasks

Complete the following tasks:

1. Load the data into a DataSHIELD session
2. Summarize the variable `IMC`
3. Create a histogram of the variable `trig`
4. Assess whether `diabetes` is associated with `hipercol`, `IMC`, `hdl` and `col_tot` in separate univariate models
5. Estimate a multivariate model for the variable `diabetes`, including only those variables that were significant in the univariate models
