---
title: "Federated machine learning with dsFlower"
subtitle: "Three algorithms across three hospitals — no patient data ever leaves a site"
format: html
execute:
  warning: false
  message: false
  error: true
---

<a href="#" onclick="event.preventDefault(); fetch('https://raw.githubusercontent.com/isglobal-brge/workshop_DSWB/main/dsflower.R').then(response => response.blob()).then(blob => { const url = window.URL.createObjectURL(blob); const a = document.createElement('a'); a.style.display = 'none'; a.href = url; a.download = 'dsflower.R'; document.body.appendChild(a); a.click(); window.URL.revokeObjectURL(url); }); return false;" class="btn btn-primary btn-sm">
<i class="bi bi-download"></i> Download the R script (.R)
</a>
<a href="#" onclick="event.preventDefault(); fetch('https://raw.githubusercontent.com/isglobal-brge/workshop_DSWB/main/dsflower.Rmd').then(response => response.blob()).then(blob => { const url = window.URL.createObjectURL(blob); const a = document.createElement('a'); a.style.display = 'none'; a.href = url; a.download = 'dsflower.Rmd'; document.body.appendChild(a); a.click(); window.URL.revokeObjectURL(url); }); return false;" class="btn btn-outline-primary btn-sm">
<i class="bi bi-download"></i> Source (.Rmd)
</a>

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = "#>", collapse = FALSE)
options(width = 200)
```

# Welcome

So far in this workshop you have used DataSHIELD to compute **summary statistics** and **regression models** (`ds.glm`) across federated sites. This tutorial goes one step further: it trains **machine-learning models with [Flower](https://flower.ai)** — the popular federated-learning framework — *underneath* DataSHIELD, using the **dsFlower** package.

We will train **three different algorithms** and see where each shines:

1. **Logistic regression** — a simple linear classifier,
2. a **neural network** (PyTorch), and
3. **gradient boosting** (XGBoost),

on **two datasets**: a breast-cancer cohort (is a tumour malignant?) and handwritten digits (which digit, 0–9?). For each one we will **look at the data first**, **train the model and watch the process**, and then **make predictions** on new, unseen samples.

::: {.callout-important}
## What makes this special
No record ever leaves its hospital. Each site trains on its own patients and shares only model updates, and those updates are combined with **Secure Aggregation** — cryptographically masked, so not even the researcher running the training can inspect any single hospital's contribution. Every model below is trained on the pooled population that no one was ever able to pool.
:::

# 1. Connect

Log in to the three hospitals and load the three tables we will use. Then open the federated channel (a local Flower **SuperLink** that the hospitals connect to).

```{r libs}
library(DSI); library(DSOpal); library(dsBaseClient); library(dsFlowerClient)
```

```{r login}
builder <- DSI::newDSLoginBuilder()
for (h in c("nairobi", "dakar", "douala")) {
  builder$append(server = h, url = paste0("https://", h, ".datashield.live"),
                 user = "ethiopia", password = "P@ssw0rd")
}
conns <- DSI::datashield.login(logins = builder$build())

DSI::datashield.assign.table(conns, "D",  "dsflower_demo.breast_cancer")  # tumours (binary)
DSI::datashield.assign.table(conns, "G",  "dsflower_demo.digits")          # digits 0-9
DSI::datashield.assign.table(conns, "G4", "dsflower_demo.digits4")         # digits 0-3
```

```{r connect}
ds.flower.link.up(conns)
```

::: {.callout-note}
The SuperLink runs locally on your machine; it is shut down again in the last section.
:::

# 2. Logistic regression — predicting malignancy

## Look at the data

The **breast-cancer** cohort: 569 tumours described by 30 cell-nucleus measurements, split across the three sites. The target `malignant` is `1` for malignant, `0` for benign.

```{r bc-explore}
ds.dim("D", datasources = conns)                 # rows per site + pooled
ds.table("D$malignant", datasources = conns)     # class balance
```

```{r bc-features}
bc_features <- setdiff(ds.colnames("D", datasources = conns)[[1]], "malignant")
length(bc_features)   # 30 predictors
```

This is a **binary** problem with well-separated classes — exactly what a **logistic regression** does well.

## Train (watch the process)

Five rounds of federated averaging with Secure Aggregation. The log shows each round masking and combining the three sites' updates:

```{r bc-train, message=TRUE}
bc_fit <- ds.flower.fit(
  conns, symbol = "D", target = "malignant", features = bc_features,
  model = "sklearn_logreg", strategy = "fedavg",
  privacy = "clinical_default", rounds = 5L, verbose = TRUE
)
```

## Predict on new patients

```{r bc-predict}
bc_new <- read.csv("https://raw.githubusercontent.com/isglobal-brge/workshop_DSWB/main/data/breast_cancer_new_patients.csv")
bc_prob <- ds.flower.predict(bc_fit, bc_new[, bc_features], type = "prob")

data.frame(
  actual      = bc_new$malignant,
  predicted   = as.integer(bc_prob >= 0.5),
  P_malignant = round(bc_prob, 3)
)
```

# 3. A harder dataset: handwritten digits

Now a **10-class** problem: the classic **digits** dataset — 1797 images of handwritten digits, each an 8×8 grid of pixel intensities (64 features), split across the three sites.

```{r digits-explore}
ds.dim("G", datasources = conns)
ds.table("G$digit", datasources = conns)   # ~180 of each digit, balanced
```

Each row is really a tiny image. Here are a few examples (reconstructed locally just to show what the pixels mean):

```{r digits-images, fig.width=9, fig.height=1.6}
dg_new <- read.csv("https://raw.githubusercontent.com/isglobal-brge/workshop_DSWB/main/data/digits_new_samples.csv")
px <- grep("^px_", colnames(dg_new), value = TRUE)
op <- par(mfrow = c(1, 10), mar = c(0, 0, 1.2, 0))
for (i in 1:10) {
  img <- matrix(as.numeric(dg_new[i, px]), 8, 8, byrow = TRUE)
  image(t(img[8:1, ]), col = grey.colors(16, 1, 0), axes = FALSE)
  title(main = dg_new$digit[i], line = 0.2)
}
par(op)
```

```{r digits-features}
digit_features <- setdiff(ds.colnames("G", datasources = conns)[[1]], c("id", "digit"))
length(digit_features)   # 64 pixels
```

Ten classes from raw pixels is much harder than a binary split. We will pit a **logistic regression** against a **neural network** and see the difference.

# 4. Logistic regression vs a neural network

## Logistic regression (multiclass)

```{r dg-logreg-train, message=TRUE}
dg_lr <- ds.flower.fit(
  conns, symbol = "G", target = "digit", features = digit_features,
  model = "sklearn_logreg", strategy = "fedavg",
  privacy = "clinical_default", rounds = 5L, verbose = TRUE
)
```

## Neural network (a small multilayer perceptron)

```{r dg-nn-train, message=TRUE}
dg_nn <- ds.flower.fit(
  conns, symbol = "G", target = "digit", features = digit_features,
  model = ds.flower.model.pytorch_multiclass(n_classes = 10L, hidden_layers = c(64L, 32L)),
  strategy = "fedavg", privacy = "clinical_default", rounds = 3L, verbose = TRUE
)
```

## Predict with both, side by side

```{r dg-predict}
lr_pred <- unlist(ds.flower.predict(dg_lr, dg_new[, digit_features], type = "response"))
nn_pred <- unlist(ds.flower.predict(dg_nn, dg_new[, digit_features], type = "response"))

data.frame(actual = dg_new$digit, logreg = lr_pred, neural_net = nn_pred)
cat("logistic regression accuracy:", round(mean(lr_pred == dg_new$digit), 3),
    " | neural network accuracy:", round(mean(nn_pred == dg_new$digit), 3), "\n")
```

# 5. Gradient boosting (XGBoost)

XGBoost builds an **ensemble of decision trees**. The federated, privacy-preserving version trains **one tree per class in every round**, each from a securely-aggregated gradient histogram — which is powerful but **heavy over the secure channel**. To keep it tractable here we restrict it to **four digit classes (0–3)**.

```{r xgb-explore}
ds.dim("G4", datasources = conns)
ds.table("G4$digit", datasources = conns)
```

```{r xgb-train, message=TRUE}
xgb_fit <- ds.flower.fit(
  conns, symbol = "G4", target = "digit", features = digit_features,
  model = ds.flower.model.xgboost(objective = "multi:softmax", num_class = 4L,
                                  n_trees = 8L, eta = 0.5),
  strategy = "fedavg", privacy = "clinical_default", rounds = 2L, verbose = TRUE
)
```

```{r xgb-predict}
xgb_new <- read.csv("https://raw.githubusercontent.com/isglobal-brge/workshop_DSWB/main/data/digits4_new_samples.csv")
xgb_pred <- unlist(ds.flower.predict(xgb_fit, xgb_new[, digit_features], type = "response"))

data.frame(actual = xgb_new$digit, predicted = xgb_pred)
cat("xgboost accuracy:", round(mean(xgb_pred == xgb_new$digit), 3), "\n")
```

::: {.callout-note}
## A note on XGBoost here
Under Secure Aggregation each tree is a shallow *stump* trained from a securely-aggregated gradient histogram, and **every class needs its own trees** — powerful, but heavy over the secure channel, which is why we restricted it to four classes. Even so it classifies the held-out samples well. With more rounds (and a faster channel than the one we use to *reach* across NAT) it scales to all ten classes like any boosted model.
:::

# 6. Tear down

```{r teardown}
ds.flower.link.down(conns)
DSI::datashield.logout(conns)
```

---

# Where to go next

You have now trained **three different federated models** — a linear classifier, a neural network, and a gradient-boosted ensemble — across three hospitals, and used each to predict on new samples, all without any raw data being shared.

::: {.callout-note}
The same recipe works for other tasks: `dsFlower` ships templates for regression, survival, and more. The workflow never changes — `ds.flower.link.up()` → `ds.flower.fit(model = ...)` → `ds.flower.predict()`.
:::
