---
title: "Federated machine learning with dsFlower"
subtitle: "Train a real model across three hospitals — without any data leaving its site, and without opening a single port"
format: html
execute:
  warning: false
  message: false
  error: true
---

<a href="#" onclick="event.preventDefault(); fetch('https://raw.githubusercontent.com/isglobal-brge/workshop_DSWB/main/dsflower.R').then(response => response.blob()).then(blob => { const url = window.URL.createObjectURL(blob); const a = document.createElement('a'); a.style.display = 'none'; a.href = url; a.download = 'dsflower.R'; document.body.appendChild(a); a.click(); window.URL.revokeObjectURL(url); }); return false;" class="btn btn-primary btn-sm">
<i class="bi bi-download"></i> Download the R script (.R)
</a>
<a href="#" onclick="event.preventDefault(); fetch('https://raw.githubusercontent.com/isglobal-brge/workshop_DSWB/main/dsflower.Rmd').then(response => response.blob()).then(blob => { const url = window.URL.createObjectURL(blob); const a = document.createElement('a'); a.style.display = 'none'; a.href = url; a.download = 'dsflower.Rmd'; document.body.appendChild(a); a.click(); window.URL.revokeObjectURL(url); }); return false;" class="btn btn-outline-primary btn-sm">
<i class="bi bi-download"></i> Source (.Rmd)
</a>

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = "#>", collapse = FALSE, eval = FALSE)
options(width = 200)
```

# Welcome

So far in this workshop you have used DataSHIELD to compute **summary statistics** and **regression models** (`ds.glm`) across federated sites. Those methods send a formula to each server, which returns aggregated numbers.

This tutorial does something different: it trains a **machine-learning model with [Flower](https://flower.ai)** — the popular federated-learning framework — *underneath* DataSHIELD, using the **dsFlower** package. Each hospital trains a logistic-regression model **on its own patients**, and only the model weights are combined. We will:

1. connect to the three sites and load a breast-cancer dataset,
2. look at what's in the data,
3. **train a model federated across all three hospitals**, and
4. **make predictions** for new patients.

::: {.callout-important}
## What makes this special
The model is trained with **Secure Aggregation** (SecAgg+): each site's weight update is cryptographically masked, so not even the researcher running the training can see any individual hospital's contribution — only the final averaged model. And the whole thing runs **over Tor**, so the federation works even when the hospital servers are behind a firewall/NAT and cannot expose any public address. No VPN, no port-forwarding, no public IP.
:::

::: {.callout-note}
## This is a presenter-led demo
Unlike the other tutorials, the code below is **not executed when this page is built** — federated training needs a Flower *SuperLink* and Tor running **on your own laptop** (started automatically by `ds.flower.tor.up()`), plus the `dsFlowerClient` package. The outputs shown are the **real results** of a live run against `nairobi`, `dakar` and `douala`. To run it yourself, install `dsFlowerClient` and execute the downloaded `.R` script.
:::

# 1. Connect and load the data

The dataset is the classic **Wisconsin breast-cancer** cohort (569 patients), split across the three hospitals. Each row is a tumour with 30 cell-nucleus measurements; the target `malignant` is `1` for malignant, `0` for benign.

```{r libs}
library(DSI); library(DSOpal); library(dsBaseClient); library(dsFlowerClient)
```

```{r login}
builder <- DSI::newDSLoginBuilder()
builder$append(server = "nairobi", url = "https://nairobi.datashield.live",
               user = "ethiopia", password = "P@ssw0rd",
               table = "dsflower_demo.breast_cancer")
builder$append(server = "dakar",   url = "https://dakar.datashield.live",
               user = "ethiopia", password = "P@ssw0rd",
               table = "dsflower_demo.breast_cancer")
builder$append(server = "douala",  url = "https://douala.datashield.live",
               user = "ethiopia", password = "P@ssw0rd",
               table = "dsflower_demo.breast_cancer")
conns <- DSI::datashield.login(logins = builder$build(), assign = TRUE, symbol = "D")
```

The table is now loaded into the symbol `D` on every server.

# 2. What's in the data?

Always look before you model. How many patients does each site hold?

```{r dims}
ds.dim("D", datasources = conns)
```

```
$`dimensions of D in nairobi`
[1] 190  31
$`dimensions of D in dakar`
[1] 190  31
$`dimensions of D in douala`
[1] 189  31
$`dimensions of D in combined studies`
[1] 569  31
```

569 patients in total — but no site sees more than its own ~190. There are 30 feature columns plus the `malignant` target:

```{r features}
features <- setdiff(ds.colnames("D", datasources = conns)[[1]], "malignant")
length(features)   # 30 predictors
```

And the outcome we want to predict — how many benign vs malignant, pooled across all three hospitals?

```{r outcome}
ds.table("D$malignant", datasources = conns)
```

```
D$malignant
  0   1
357 212
```

357 benign and 212 malignant — a healthy class balance to learn from.

# 3. Bring the federation to your laptop (over Tor)

A Flower model needs a central **SuperLink** that the hospital **SuperNodes** dial into. Normally that means *you* must expose a public server — impossible from a laptop behind NAT. `dsFlower` solves this by publishing your SuperLink as a **Tor hidden service**: each hospital connects out through Tor, so nothing needs a public address.

One call starts a SuperLink on your machine, publishes it as a `.onion`, and brings all three nodes onto Tor:

```{r torup}
ds.flower.tor.up(conns)
```

```
SuperLink started (PID: 99435)
  Fleet API (SuperNodes): 127.0.0.1:9092
  Control API (flwr run): 127.0.0.1:9093
Publishing SuperLink as a Tor hidden service (bootstrapping Tor)...
  SuperLink onion: kx3ffiojuk625h5e2ouyiulw3g3cehp36jnxcrxisqqoqkcbu5mevoqd.onion:9092
Bringing nodes onto Tor...
Tor transport ready.
```

::: {.callout-tip}
## What just happened
Your laptop is now reachable by the hospitals at a `.onion` address that exists only for this session. Tor trades a little speed for the ability to federate with **zero infrastructure** — no VPN, no cloud host, no firewall changes. When you tear it down (Section 6) the onion disappears.
:::

# 4. Train the model — federated, with Secure Aggregation

Now the actual training. We fit a federated **logistic regression** (`sklearn_logreg`) with the **FedAvg** strategy for 5 rounds, under the `clinical_default` privacy profile (which enforces Secure Aggregation):

```{r fit}
fit <- ds.flower.fit(
  conns,
  symbol   = "D",
  target   = "malignant",
  features = features,
  model    = "sklearn_logreg",
  strategy = "fedavg",
  privacy  = "clinical_default",   # enforces Secure Aggregation (SecAgg+)
  rounds   = 5L,
  verbose  = TRUE
)
```

```
[ROUND 1] Secure aggregation commencing. ... sampled 3 clients (out of 3)
          aggregate_fit: received 3 results and 0 failures
          Secure aggregation completed.
...
[ROUND 5] Secure aggregation completed.
[SUMMARY] Run finished 5 round(s) in 236.89s
Model saved to ./dsflower_output/sklearn_logreg_FedAvg_5r_...
```

::: {.callout-note}
## Read the log
Every round shows `Secure aggregation commencing` → `completed`, and `received 3 results and 0 failures` — all three hospitals contributed, and each update was masked so no individual contribution was ever visible. The model that comes back is the **averaged** logistic regression, trained on all 569 patients without any of them leaving their hospital.
:::

# 5. Make predictions for new patients

The trained model is an ordinary object you can predict with. Here are six **new patients** (three benign, three malignant) the model has never seen:

```{r predict}
new_patients <- read.csv("https://raw.githubusercontent.com/isglobal-brge/workshop_DSWB/main/data/breast_cancer_new_patients.csv")

prob <- ds.flower.predict(fit, new_patients[, features], type = "prob")

data.frame(
  actual      = new_patients$malignant,
  predicted   = as.integer(prob >= 0.5),
  P_malignant = round(prob, 3)
)
```

```
  actual predicted P_malignant
1      0         0       0.025
2      0         0       0.000
3      0         0       0.004
4      1         1       1.000
5      1         1       1.000
6      1         1       1.000
```

All six correct — the federated model cleanly separates benign from malignant, with calibrated probabilities. **A model trained across three hospitals, predicting for a new patient, with no raw data ever shared.**

# 6. Tear down

Stop the SuperNodes, remove the Tor onion, and log out:

```{r teardown}
ds.flower.tor.down(conns)
DSI::datashield.logout(conns)
```

---

# Where to go next

You have now trained a **federated machine-learning model** and used it to predict — the same DataSHIELD principle (data never moves), extended from summary statistics to full ML via Flower.

::: {.callout-tip}
## A note on speed and scale
Tor makes the federation work from anywhere, but it is **not fast** — this 5-round run took ~4 minutes, most of it network latency. For a tiny logistic regression that overhead dominates; for larger models on a trusted network you can swap the transport for a direct or coordinator-based one. The point of Tor here is **reach**: training across locked-down hospital servers that can't expose anything.
:::

::: {.callout-note}
## Other models
`dsFlower` ships templates beyond logistic regression — `pytorch_*` neural networks (with patient-level differential privacy via Opacus) and `xgboost`. The workflow is identical: `ds.flower.tor.up()` → `ds.flower.fit(model = ...)` → `ds.flower.predict()`.
:::
