---
title: "Federated OMOP analysis with dsOMOP"
subtitle: "A hands-on tutorial — exploring real hospital data across three sites, without the data ever leaving them"
format: html
execute:
  warning: false
  message: false
  error: true
---

<a href="#" onclick="event.preventDefault(); fetch('https://raw.githubusercontent.com/isglobal-brge/workshop_DSWB/main/dsomop.R').then(response => response.blob()).then(blob => { const url = window.URL.createObjectURL(blob); const a = document.createElement('a'); a.style.display = 'none'; a.href = url; a.download = 'dsomop.R'; document.body.appendChild(a); a.click(); window.URL.revokeObjectURL(url); }); return false;" class="btn btn-primary btn-sm">
<i class="bi bi-download"></i> Download the R script (.R)
</a>
<a href="#" onclick="event.preventDefault(); fetch('https://raw.githubusercontent.com/isglobal-brge/workshop_DSWB/main/dsomop.Rmd').then(response => response.blob()).then(blob => { const url = window.URL.createObjectURL(blob); const a = document.createElement('a'); a.style.display = 'none'; a.href = url; a.download = 'dsomop.Rmd'; document.body.appendChild(a); a.click(); window.URL.revokeObjectURL(url); }); return false;" class="btn btn-outline-primary btn-sm">
<i class="bi bi-download"></i> Source (.Rmd)
</a>

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = "#>", collapse = FALSE)
```

# Welcome

In this tutorial you will analyse **real-world hospital data** that lives in **three different hospitals at once** — and you will do it **without ever seeing or downloading a single patient's record**.

- **OMOP CDM** is a standard *shape* for health data: every hospital stores patients, diagnoses, drugs and labs in the same tables (`person`, `condition_occurrence`, `drug_exposure`, `measurement`, ...).
- **DataSHIELD** is the safety layer: you send *commands*, each hospital runs them on its own data and returns **only aggregated, non-disclosive results**.
- **dsOMOP** is the bridge that lets DataSHIELD speak OMOP.

::: {.callout-important}
## The golden rule
You never hold patient-level data on your laptop. Every result is checked by a **disclosure filter** on each server before it is returned. If a result would describe too few people, the server hides it.
:::

We work with **MIMIC** (an intensive-care database, ~100 patients here): connect → explore → learn the **recipe** for building analysis tables → filter → and finish with a real **regression model**.

# 1. Connecting to the federation

```{r libs}
library(DSI)
library(DSOpal)
library(dsBaseClient)
library(dsOMOPClient)
```

```{r login}
builder <- DSI::newDSLoginBuilder()
builder$append(server = "nairobi", url = "https://nairobi.datashield.live",
               user = "ethiopia", password = "P@ssw0rd", profile = "omop")
builder$append(server = "douala",  url = "https://douala.datashield.live",
               user = "ethiopia", password = "P@ssw0rd", profile = "omop")
builder$append(server = "dakar",   url = "https://dakar.datashield.live",
               user = "ethiopia", password = "P@ssw0rd", profile = "omop")
conns <- DSI::datashield.login(logins = builder$build())

ds.omop.connect(resource = "omop_demo.mimic", symbol = "omop", conns = conns)
```

::: {.callout-warning}
Forgetting `profile = "omop"` is the #1 beginner mistake — without it the `ds.omop.*` functions aren't available.
:::

# 2. Reading the map: tables and columns

The clinical (`CDM`) tables (consistent across the three sites, so one view represents the federation):

```{r tables}
tabs <- ds.omop.tables(symbol = "omop", conns = conns)[[1]]
tabs[tabs$schema_category == "CDM", c("table_name", "has_person_id")]
```

The columns of the `measurement` table (where labs and vital signs live):

```{r columns}
ds.omop.columns("measurement", symbol = "omop", conns = conns)[[1]]
```

# 3. Who are these patients?

Gender, **pooled** across the three sites:

```{r gender}
ds.omop.concept.prevalence("person", concept_col = "gender_concept_id",
                           scope = "pooled", symbol = "omop", conns = conns)$pooled
```

::: {.callout-important}
## Concepts translate themselves
You got `MALE` / `FEMALE`, not codes like `8507` / `8532` — any `*_concept_id` column is automatically translated. **Concept translation is on by default.**
:::

And their **birth years** — `ds.omop.value.histogram()` draws a disclosure-safe, federation-wide histogram of any numeric column when you pass `plot = TRUE`:

```{r birth-hist, fig.width=7, fig.height=4}
ds.omop.value.histogram("person", value_col = "year_of_birth", nbins = 6, plot = TRUE,
                        xlab = "Year of birth", main = "Patient birth years (pooled)",
                        symbol = "omop", conns = conns)
```

::: {.callout-note}
This extract was re-anchored so its records run through **2024**; subtract a birth year from 2024 for an age (earlier birth years = older patients). We'll get age directly from the recipe in Section 7.
:::

# 4. Discovering what was recorded

The most common diagnoses, **pooled** and ranked by distinct patients:

```{r topcond}
ds.omop.concept.prevalence("condition_occurrence", metric = "persons",
                           top_n = 10, scope = "pooled", symbol = "omop", conns = conns)$pooled
```

Don't know a concept's code? **Search** for it:

```{r search}
ds.omop.concept.search("hyperlipidemia", domain = "Condition",
                       limit = 5, symbol = "omop", conns = conns)$pooled
```

And **look up** codes you already have:

```{r lookup}
ds.omop.concept.lookup(c(320128, 432867, 3027018), symbol = "omop", conns = conns)$pooled
```

# 5. Numeric measurements: shape matters

`ds.omop.column.stats()` returns a disclosure-safe summary (counts, mean, sd, a 5th–95th-percentile band — never the true min/max). Heart rate (`3027018`):

```{r hr-stats}
ds.omop.column.stats("measurement", "value_as_number", concept_id = 3027018,
                     scope = "pooled", symbol = "omop", conns = conns)$pooled
```

Its distribution — again with `plot = TRUE`:

```{r hr-hist, fig.width=7, fig.height=4}
ds.omop.value.histogram("measurement", value_col = "value_as_number", concept_id = 3027018,
                        nbins = 7, plot = TRUE, xlab = "Heart rate (bpm)",
                        main = "Heart rate (pooled across sites)",
                        symbol = "omop", conns = conns)
```

Heart rate is **roughly symmetric**. Contrast it with **serum creatinine** (`3016723`):

```{r creat-hist, fig.width=7, fig.height=4}
ds.omop.value.histogram("measurement", value_col = "value_as_number", concept_id = 3016723,
                        nbins = 9, plot = TRUE, col = "#C44E52", xlab = "Creatinine (mg/dL)",
                        main = "Serum creatinine (pooled across sites)",
                        symbol = "omop", conns = conns)
```

::: {.callout-important}
## Symmetric vs skewed
Creatinine has a **long right tail** (a minority with kidney injury) that pulls its mean above the median — so the mean overstates the "typical" patient. Always look at the shape before trusting a mean.
:::

# 6. A categorical measurement: heart rhythm

Some measurements store a *category* in `value_as_concept_id` instead of a number. Heart **rhythm** is the measurement concept `3022318`; its recorded answers live in `value_as_concept_id`. Count them, pooled:

```{r rhythm}
ds.omop.value.counts("measurement", "value_as_concept_id", concept_id = 3022318,
                     scope = "pooled", symbol = "omop", conns = conns)$pooled
```

::: {.callout-important}
## How to read this
- The first argument (`"measurement"`) is the **table** we look in; `concept_id = 3022318` is the measurement; `value_as_concept_id` is the column we count. (For a social fact you'd look in `"observation"` instead.)
- **`concept_name`** is each answer's translated label; **`value`** is its raw concept code (e.g. `45877096` = "Sinus rhythm").
- **`concept_0`** (value `0`) is OMOP's **unmapped** bucket — records whose value didn't map to a standard concept. It is *not* suppression and not a bug.
:::

# 7. The recipe: building analysis tables

To **model** anything we need one row per patient and one column per variable. You describe that table with a **recipe**; the servers build it for you. Every recipe has the same shape — here is the skeleton (don't run this; it's a template):

```{r recipe-skeleton, eval=FALSE}
omop_recipe(
  variables = list(    # WHAT columns you want
    omop_variable(),       # a person column, or something built from a concept_id
    omop_variable_age(),   # the patient's age
    omop_variable()        # ...add as many as you need
  ),
  filters = list(),    # OPTIONAL: restrict the population (Section 8)
  output  = omop_output()  # the table SHAPE you want (usually wide = one row/patient)
)
recipe_execute()       # build it on each server into a named symbol
```

## 7.1 Start simple — sex and age

```{r recipe-simple}
rec0 <- omop_recipe(
  variables = list(
    omop_variable(table = "person", column = "gender_concept_id", format = "sex_mf", name = "sex"),
    omop_variable_age(name = "age", year = 2024)
  ),
  output = omop_output(name = "study", type = "wide")
)
recipe_execute(rec0, out = c(study = "M0"), symbol = "omop", conns = conns)
```

**Always inspect what the server built**, then look at each variable:

```{r ratify-simple}
ds.colnames("M0", datasources = conns)
ds.dim("M0", datasources = conns)
```

```{r m0-sex}
ds.table("M0$sex", datasources = conns)$output.list[["TABLES.COMBINED_all.sources_counts"]]
```

```{r m0-age}
ds.summary("M0$age", datasources = conns)[[1]][["quantiles & mean"]]
```

::: {.callout-note}
## Why `year = 2024`?
Age needs a reference year. This extract runs through 2024, so we anchor age there rather than to "today" (which would drift on every rebuild). The age **filter** in Section 8 takes the same `year =`.
:::

## 7.2 Grow it — numbers, flags and counts

Mix the variable types we met: a categorical (`sex`), a numeric mean (`heart_rate`), two **binary** condition flags, and a **count** (`n_visits`):

```{r recipe-rich}
rec <- omop_recipe(
  variables = list(
    omop_variable(table = "person", column = "gender_concept_id", format = "sex_mf", name = "sex"),
    omop_variable_age(name = "age", year = 2024),
    omop_variable(table = "measurement", concept_id = 3027018, format = "mean", name = "heart_rate"),
    omop_variable(table = "condition_occurrence", concept_id = 320128, format = "binary", name = "hypertension"),
    omop_variable(table = "condition_occurrence", concept_id = 432867, format = "binary", name = "hyperlipidemia"),
    omop_variable(table = "visit_occurrence", format = "count", name = "n_visits")
  ),
  output = omop_output(name = "study", type = "wide")
)
recipe_execute(rec, out = c(study = "M"), symbol = "omop", conns = conns)
ds.colnames("M", datasources = conns)
ds.dim("M", datasources = conns)
```

**Look at every variable in its natural way.** Numeric columns → a quantile summary:

```{r check-num}
ds.summary("M$age",        datasources = conns)[[1]][["quantiles & mean"]]
ds.summary("M$heart_rate", datasources = conns)[[1]][["quantiles & mean"]]
ds.summary("M$n_visits",   datasources = conns)[[1]][["quantiles & mean"]]
```

Categorical / binary columns → a frequency table:

```{r check-sex}
ds.table("M$sex",            datasources = conns)$output.list[["TABLES.COMBINED_all.sources_counts"]]
```

```{r check-htn}
ds.table("M$hypertension",   datasources = conns)$output.list[["TABLES.COMBINED_all.sources_counts"]]
```

```{r check-hyperlip}
ds.table("M$hyperlipidemia", datasources = conns)$output.list[["TABLES.COMBINED_all.sources_counts"]]
```

::: {.callout-note}
## `binary` vs `count`
`format = "binary"` is *does this patient have it at all?* (0/1). `format = "count"` is *how many records?* — here `n_visits` is right-skewed (most patients few visits, a handful many).
:::

## 7.3 One recipe, several tables

A recipe can declare **several outputs**, each selecting a **different subset of variables**, so one extraction produces several purpose-built tables (e.g. a demographics table and a clinical one) — list them with `outputs = list(...)`:

```{r recipe-multi}
rec_multi <- omop_recipe(
  variables = list(
    omop_variable(table = "person", column = "gender_concept_id", format = "sex_mf", name = "sex"),
    omop_variable_age(name = "age", year = 2024),
    omop_variable(table = "measurement", concept_id = 3027018, format = "mean", name = "heart_rate"),
    omop_variable(table = "condition_occurrence", concept_id = 320128, format = "binary", name = "hypertension")
  ),
  outputs = list(
    omop_output(name = "demographics", variables = c("sex", "age"), type = "wide"),
    omop_output(name = "clinical",     variables = c("heart_rate", "hypertension"), type = "wide")
  )
)
recipe_execute(rec_multi, out = c(demographics = "DEMO", clinical = "CLIN"),
               symbol = "omop", conns = conns)
ds.colnames("DEMO", datasources = conns)
ds.colnames("CLIN", datasources = conns)
```

# 8. Filtering at extraction

Often you only want a **subpopulation**. Add a `filters` element so the servers only build the rows you ask for. Filters can only ever **shrink** a population, so they are safe. Here: **patients aged 50+ who have hyperlipidemia**:

```{r recipe-filtered}
rec_sub <- omop_recipe(
  variables = list(
    omop_variable(table = "person", column = "gender_concept_id", format = "sex_mf", name = "sex"),
    omop_variable_age(name = "age", year = 2024)
  ),
  filters = list(
    older       = omop_filter_age(min = 50, year = 2024),
    with_lipids = omop_filter_has_concept(concept_id = 432867, table = "condition_occurrence")
  ),
  output = omop_output(name = "study", type = "wide")
)
recipe_execute(rec_sub, out = c(study = "SUB"), symbol = "omop", conns = conns)
```

The cohort is smaller, and the minimum age respects the 50+ floor:

```{r filtered-check}
ds.dim("SUB", datasources = conns)
ds.summary("SUB$age", datasources = conns)[[1]][["quantiles & mean"]]
```

::: {.callout-note}
The safe filter types all shrink a population: `omop_filter_sex()`, `omop_filter_age()` / age groups, `omop_filter_has_concept()`, and ranges. (You can also restrict by sex with `omop_filter_sex("F")`.)
:::

# 9. A real model: what predicts a diagnosis?

A logistic regression of **hyperlipidemia** on age and sex, fitted jointly across the three hospitals. We show the raw `ds.glm` coefficients (`P_OR` is the odds ratio, with its 95% CI in the `*.P_OR` columns):

```{r glm1}
fit1 <- ds.glm(formula = "M$hyperlipidemia ~ M$age + M$sex",
               family = "binomial", datasources = conns)
fit1$coefficients
```

::: {.callout-important}
## Reading it
**age** is a clear, **significant** predictor (`P_OR` > 1, tiny `p-value`); **sex** is not (`P_OR` ≈ 1). One clear effect, one null — produced without touching a patient record.
:::

A **multivariable** model — hypertension on age, sex and hyperlipidemia:

```{r glm2}
fit2 <- ds.glm(formula = "M$hypertension ~ M$age + M$sex + M$hyperlipidemia",
               family = "binomial", datasources = conns)
fit2$coefficients
```

::: {.callout-note}
Age stays significant; sex is borderline; the comorbidity adjustment is not significant. On ~100 patients power is limited — the larger **GiBleed** study (next) delivers a strongly significant result.
:::

```{r logout}
DSI::datashield.logout(conns)
```

# Where to go next

You have the full toolkit: **connect → explore → recipe (variables / format / filters / output, single or multiple) → execute → ratify → summarise → model**. Next: the **[exercise sheet](omop-exercises.html)**, then the larger **[GiBleed study](gibleed.html)**.
