---
title: "Federated OMOP analysis with dsOMOP"
subtitle: "A hands-on tutorial — exploring real hospital data across three sites, without the data ever leaving them"
format:
  html:
    df-print: kable
execute:
  warning: false
  message: false
  error: true
---

<a href="#" onclick="event.preventDefault(); fetch('https://raw.githubusercontent.com/isglobal-brge/workshop_DSWB/main/dsomop.R').then(response => response.blob()).then(blob => { const url = window.URL.createObjectURL(blob); const a = document.createElement('a'); a.style.display = 'none'; a.href = url; a.download = 'dsomop.R'; document.body.appendChild(a); a.click(); window.URL.revokeObjectURL(url); }); return false;" class="btn btn-primary btn-sm">
<i class="bi bi-download"></i> Download the R script (.R)
</a>
<a href="#" onclick="event.preventDefault(); fetch('https://raw.githubusercontent.com/isglobal-brge/workshop_DSWB/main/dsomop.Rmd').then(response => response.blob()).then(blob => { const url = window.URL.createObjectURL(blob); const a = document.createElement('a'); a.style.display = 'none'; a.href = url; a.download = 'dsomop.Rmd'; document.body.appendChild(a); a.click(); window.URL.revokeObjectURL(url); }); return false;" class="btn btn-outline-primary btn-sm">
<i class="bi bi-download"></i> Download the source (.Rmd)
</a>

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = "#>", collapse = FALSE)
```

# Welcome

In this tutorial you will analyse **real-world hospital data** that lives in **three different hospitals at once** — and you will do it **without ever seeing or downloading a single patient's record**.

The pieces:

- **OMOP CDM** is a standard *shape* for health data. Every hospital stores its patients, diagnoses, drugs and lab results in the same tables (`person`, `condition_occurrence`, `drug_exposure`, `measurement`, ...). Learn it once, use it anywhere.
- **DataSHIELD** is the safety layer. You send *commands* to each hospital; the hospital runs them on its own data and returns **only aggregated, non-disclosive results**. Individual rows never travel.
- **dsOMOP** is the bridge that lets DataSHIELD speak OMOP.

::: {.callout-important}
## The golden rule
You never hold patient-level data on your laptop. Every result is checked by a **disclosure filter** on each server before it is returned. If a result would describe too few people, the server hides it. This explains much of what you'll see.
:::

We work with **MIMIC** (an intensive-care database, ~100 patients here) throughout. The plan, step by step:

1. **Connect** to the three sites.
2. **Explore** — what tables, patients, conditions and measurements exist?
3. Learn the **recipe** — how to ask the servers to build an analysis table, growing from a one-line recipe to a richer one.
4. **Filter** at extraction (by condition, sex, age).
5. Fit a real **regression model** and read the result.

---

# 1. Connecting to the federation

```{r libs}
library(DSI)
library(DSOpal)
library(dsBaseClient)
library(dsOMOPClient)
```

Describe the three servers and log in (same training login on each, asking for the **`omop`** profile), then attach the MIMIC database under the symbol `"omop"`:

```{r login}
builder <- DSI::newDSLoginBuilder()
builder$append(server = "nairobi", url = "https://nairobi.datashield.live",
               user = "ethiopia", password = "P@ssw0rd", profile = "omop")
builder$append(server = "douala",  url = "https://douala.datashield.live",
               user = "ethiopia", password = "P@ssw0rd", profile = "omop")
builder$append(server = "dakar",   url = "https://dakar.datashield.live",
               user = "ethiopia", password = "P@ssw0rd", profile = "omop")
conns <- DSI::datashield.login(logins = builder$build())

ds.omop.connect(resource = "omop_demo.mimic", symbol = "omop", conns = conns)
```

::: {.callout-warning}
Forgetting `profile = "omop"` is the #1 beginner mistake — without it the `ds.omop.*` functions aren't available. If that happens, log out and rebuild the login with the profile set.
:::

---

# 2. Reading the map: tables and columns

dsOMOP reads each site's schema **separately** — a real federation could have different tables on different sites. Here the three sites are consistent, so one view represents the whole federation. These are the clinical (`CDM`) tables:

```{r tables}
tabs <- ds.omop.tables(symbol = "omop", conns = conns)[[1]]
tabs[tabs$schema_category == "CDM", c("table_name", "has_person_id")]
```

The columns of a table are the same across the three sites, so one view is enough. The `measurement` table is where labs and vital signs live:

```{r columns}
ds.omop.columns("measurement", symbol = "omop", conns = conns)[[1]]
```

::: {.callout-tip collapse="true"}
## Where would you look for...?
diagnoses → `condition_occurrence`; medications → `drug_exposure`; lab values / vital signs → `measurement`; social facts → `observation`; hospital/ICU stays → `visit_occurrence`. Every one has a `person_id` linking a patient's records.
:::

---

# 3. Who are these patients?

Gender lives in `person.gender_concept_id`. Rank its values, **pooled** across all three sites:

```{r gender}
ds.omop.concept.prevalence("person", concept_col = "gender_concept_id",
                           scope = "pooled", symbol = "omop", conns = conns)$pooled
```

::: {.callout-important}
## Concepts translate themselves
You got `MALE` / `FEMALE`, not codes like `8507` / `8532`. Any `*_concept_id` column is automatically translated to readable labels — **concept translation is on by default** everywhere in dsOMOP.
:::

::: {.callout-note}
## A real-world aside: de-identification
Intensive-care data is de-identified by **shifting dates**. For this teaching dataset the dates were re-anchored so its records run through **2024**, but in real clinical data the calendar is often deliberately distorted while **ages and intervals are preserved**. The lesson: derive age from the data against a known reference year — never assume "this year" means "now". We do exactly that in Section 7.
:::

---

# 4. Discovering what was recorded

The most common diagnoses, **pooled** and ranked by distinct patients:

```{r topcond}
ds.omop.concept.prevalence("condition_occurrence", metric = "persons",
                           top_n = 10, scope = "pooled", symbol = "omop", conns = conns)$pooled
```

::: {.callout-note}
## What you are (not) seeing
Counts are **banded** (rounded) so you never read an exact size, and any condition affecting **too few patients to be safe is simply not listed** — it is dropped on the server, never shown as a flagged or blank row. Absence here means "rare or absent", and you cannot tell which: that is the disclosure filter protecting individuals.
:::

Don't know a concept's code? **Search** for it (the result is the same across sites, so we read the pooled view; we keep the most useful columns):

```{r search}
hits <- ds.omop.concept.search("hyperlipidemia", domain = "Condition",
                               limit = 5, symbol = "omop", conns = conns)$pooled
hits[, c("concept_id", "concept_name", "domain_id", "vocabulary_id")]
```

And **look up** codes you already have:

```{r lookup}
ds.omop.concept.lookup(c(320128, 432867, 3027018),
                       symbol = "omop", conns = conns)$pooled
```

::: {.callout-tip collapse="true"}
## Your turn
Find the `concept_id` for a condition that interests you with `ds.omop.concept.search(..., domain = "Condition")`.
:::

---

# 5. Numeric measurements: shape matters

Measurements store a number in `value_as_number`. `ds.omop.column.stats()` returns a **disclosure-safe** summary — counts, mean, sd, and a 5th–95th-percentile band (never the true min/max, which could identify a patient).

Start with **heart rate** (`concept_id = 3027018`):

```{r hr-stats}
ds.omop.column.stats("measurement", "value_as_number", concept_id = 3027018,
                     scope = "pooled", symbol = "omop", conns = conns)$pooled
```

A picture of the distribution. `ds.omop.value.histogram()` returns disclosure-safe **bins per site** (clipped to the 5th–95th percentile, with sparse bins dropped); we put them on a common grid and draw the federation-wide shape:

```{r hr-hist, fig.width=7, fig.height=4}
draw_hist <- function(concept_id, nbins, xlab, main, col = "#4C72B0", digits = 0) {
  h <- ds.omop.value.histogram("measurement", value_col = "value_as_number",
                               concept_id = concept_id, symbol = "omop", conns = conns)
  bins <- do.call(rbind, h$per_site)
  bins <- subset(bins, !is.na(count))
  mid  <- (bins$bin_start + bins$bin_end) / 2
  br   <- seq(min(mid), max(mid), length.out = nbins + 1)   # common equal-width bins
  grp  <- cut(mid, breaks = br, include.lowest = TRUE)      # re-bin each site onto them
  agg  <- tapply(bins$count, grp, sum); agg[is.na(agg)] <- 0
  ctr  <- round((head(br, -1) + tail(br, -1)) / 2, digits)
  barplot(as.numeric(agg), names.arg = ctr, las = 2, col = col, border = NA,
          xlab = xlab, ylab = "records", main = main)
}
draw_hist(3027018, nbins = 7, xlab = "Heart rate (bpm)",
          main = "Heart rate across the federation")
```

Heart rate is **roughly symmetric** — a single central peak, mean ≈ median. Now contrast it with a lab that is **not** symmetric, **serum creatinine** (`3016723`):

```{r creat-stats}
ds.omop.column.stats("measurement", "value_as_number", concept_id = 3016723,
                     scope = "pooled", symbol = "omop", conns = conns)$pooled
```

```{r creat-hist, fig.width=7, fig.height=4}
draw_hist(3016723, nbins = 9, xlab = "Creatinine (mg/dL)",
          main = "Serum creatinine across the federation", col = "#C44E52", digits = 1)
```

::: {.callout-important}
## Symmetric vs skewed — why it matters
Heart rate's mean sits in the middle of the bulk. Creatinine's distribution has a **long right tail** (a minority of patients with kidney injury), which **pulls the mean above the median** — so for creatinine the mean overstates the "typical" patient. Always look at the shape before trusting a mean; skewed labs are often summarised with the median, or modelled on the log scale.
:::

::: {.callout-tip collapse="true"}
## Always check the units, too
Not every measurement is clean. Body temperature (`3020891`) in this extract mixes **Celsius and Fahrenheit** records, so its histogram is bimodal and its mean (~88) is meaningless. A real federated project spends much of its time on this kind of **data-quality / unit-harmonisation** check before any analysis.
:::

---

# 6. A categorical measurement: heart rhythm

Some measurements store a *category* in `value_as_concept_id` instead of a number. Heart **rhythm** (`3022318`) is the classic example. Ask for a **pooled** federation-wide count with `scope = "pooled"`:

```{r rhythm}
ds.omop.value.counts("measurement", "value_as_concept_id",
                     concept_id = 3022318, scope = "pooled",
                     symbol = "omop", conns = conns)$pooled
```

::: {.callout-important}
## Two kinds of measurement
A **numeric** measurement (heart rate *value*) → summarise with mean / sd / histogram. A **categorical** one (heart *rhythm*) → count each level. The coded values came back as readable labels (`Sinus rhythm`, `Atrial fibrillation`, ...); the `concept_0` bucket is the **unmapped** records. Any level too rare to clear the privacy threshold is simply **absent** — never a flagged row.
:::

---

# 7. The recipe: building analysis tables

Single aggregates are great for exploring, but to **model** anything we need a table with **one row per patient** and **one column per variable**. In dsOMOP you describe that table with a **recipe**, and the servers build it for you — still without sending any rows.

A recipe has four elements:

| Element | What it is |
|---|---|
| **variables** | the columns you want — each an `omop_variable(...)`: a `person` column, or something built from a `concept_id` in an event table |
| **format** | *how* each variable is summarised per patient: `"sex_mf"`, `"binary"` (0/1 presence), `"count"` (how many records), `"mean"`, `"sd"`, ... (and `omop_variable_age()` builds age) |
| **filters** | optional restrictions applied **during** extraction (Section 8) |
| **output** | `omop_output(name, type = "wide")` — `"wide"` = one row per patient, exactly what a model needs |

## 7.1 Start simple

The smallest useful recipe: just **sex** and **age**. Preview it (a server-side dry run that builds nothing), then execute it into a server-side symbol `M0`.

```{r recipe-simple}
rec0 <- omop_recipe(
  variables = list(
    omop_variable(table = "person", column = "gender_concept_id", format = "sex_mf", name = "sex"),
    omop_variable_age(name = "age", year = 2024)
  ),
  output = omop_output(name = "study", type = "wide")
)
recipe_execute(rec0, out = c(study = "M0"), symbol = "omop", conns = conns)
```

**Always inspect what the server built** — confirm the table exists, with the columns and the number of rows you expected:

```{r ratify-simple}
ds.colnames("M0", datasources = conns)
ds.dim("M0", datasources = conns)
```

::: {.callout-note}
## Why `year = 2024`?
Age needs a reference year. This MIMIC extract runs through **2024**, so we anchor age there rather than to `today` — otherwise every rebuild of this page would silently age the patients. For your own data, set `year =` to its collection year (or use `reference = "index"` for age at cohort entry). The age **filter** in Section 8 takes the same `year =` so the two always agree.
:::

## 7.2 Grow it: numbers, flags and counts

Now mix the variable *types* we met while exploring — a categorical (`sex`), a numeric mean (`heart_rate`), two **binary** condition flags (`hypertension` 320128, `hyperlipidemia` 432867), and a **count** (`n_visits`, how many hospital visits each patient has):

```{r recipe-rich}
rec <- omop_recipe(
  variables = list(
    omop_variable(table = "person", column = "gender_concept_id", format = "sex_mf", name = "sex"),
    omop_variable_age(name = "age", year = 2024),
    omop_variable(table = "measurement", concept_id = 3027018, format = "mean", name = "heart_rate"),
    omop_variable(table = "condition_occurrence", concept_id = 320128, format = "binary", name = "hypertension"),
    omop_variable(table = "condition_occurrence", concept_id = 432867, format = "binary", name = "hyperlipidemia"),
    omop_variable(table = "visit_occurrence", format = "count", name = "n_visits")
  ),
  output = omop_output(name = "study", type = "wide")
)
recipe_execute(rec, out = c(study = "M"), symbol = "omop", conns = conns)
ds.colnames("M", datasources = conns)
ds.dim("M", datasources = conns)
```

**Inspect every variable in its natural way.** A numeric summary for the continuous columns:

```{r check-num}
ds.summary("M$heart_rate", datasources = conns)[[1]][["quantiles & mean"]]
ds.summary("M$n_visits",  datasources = conns)[[1]][["quantiles & mean"]]
```

A frequency table for the categorical / binary columns:

```{r check-sex}
as.data.frame(ds.table("M$sex", datasources = conns)$output.list[["TABLES.COMBINED_all.sources_counts"]])
```

```{r check-htn}
as.data.frame(ds.table("M$hypertension", datasources = conns)$output.list[["TABLES.COMBINED_all.sources_counts"]])
```

::: {.callout-note}
## `binary` vs `count`
`format = "binary"` answers *"does this patient have the condition at all?"* (0/1) — perfect for a yes/no risk factor. `format = "count"` answers *"how many records?"* — here, how many visits. Notice `n_visits` is **right-skewed**: most patients have a few visits, a handful have many.
:::

## 7.3 One recipe, several tables: multiple outputs

A recipe can declare **more than one output**, each selecting a **different subset of the variables** — so a single extraction can produce several purpose-built tables at once (e.g. a demographics table and a clinical table) instead of re-running the extraction. List them with `outputs = list(...)` and map each name to its own symbol:

```{r recipe-multi}
rec_multi <- omop_recipe(
  variables = list(
    omop_variable(table = "person", column = "gender_concept_id", format = "sex_mf", name = "sex"),
    omop_variable_age(name = "age", year = 2024),
    omop_variable(table = "measurement", concept_id = 3027018, format = "mean", name = "heart_rate"),
    omop_variable(table = "condition_occurrence", concept_id = 320128, format = "binary", name = "hypertension")
  ),
  outputs = list(
    omop_output(name = "demographics", variables = c("sex", "age"), type = "wide"),
    omop_output(name = "clinical",     variables = c("heart_rate", "hypertension"), type = "wide")
  )
)
recipe_execute(rec_multi, out = c(demographics = "DEMO", clinical = "CLIN"),
               symbol = "omop", conns = conns)
```

Ratify that **both** tables were built, each with only its own columns (plus the `person_id` key that lets you join them later):

```{r ratify-multi}
ds.colnames("DEMO", datasources = conns)
ds.colnames("CLIN", datasources = conns)
```

---

# 8. Filtering at extraction

Often you only want a **subpopulation**. The cleanest way is to filter **while extracting** — add a `filters` element so the servers only ever build the rows you asked for. Filters can only ever **shrink** a population (the per-patient disclosure gate still runs on the result), so they are safe.

Here we combine three filters — keep only **females**, **aged 65+**, who **have hypertension**:

```{r recipe-filtered}
rec_sub <- omop_recipe(
  variables = list(
    omop_variable(table = "person", column = "gender_concept_id", format = "sex_mf", name = "sex"),
    omop_variable_age(name = "age", year = 2024),
    omop_variable(table = "condition_occurrence", concept_id = 432867, format = "binary", name = "hyperlipidemia")
  ),
  filters = list(
    female   = omop_filter_sex("F"),
    elderly  = omop_filter_age(min = 65, year = 2024),
    with_htn = omop_filter_has_concept(concept_id = 320128, table = "condition_occurrence")
  ),
  output = omop_output(name = "study", type = "wide")
)
recipe_execute(rec_sub, out = c(study = "SUB"), symbol = "omop", conns = conns)
```

**Check the filters did what you meant** — the cohort is smaller, everyone is female, and the minimum age respects the 65+ filter:

```{r filtered-check}
ds.dim("SUB", datasources = conns)
as.data.frame(ds.table("SUB$sex", datasources = conns)$output.list[["TABLES.COMBINED_all.sources_counts"]])
ds.summary("SUB$age", datasources = conns)[[1]][["quantiles & mean"]]
```

::: {.callout-note}
These filters are just a demonstration — we built `SUB` to show how filtering narrows a cohort, but we go back to the full table `M` for the model below. The safe filter types are the ones that can only shrink a population: `omop_filter_sex()`, `omop_filter_age()` / age groups, `omop_filter_has_concept()`, and ranges; an over-narrow filter (e.g. a single birth year) is rejected so it cannot fingerprint a person.
:::

---

# 9. A real model: what predicts a diagnosis?

We have the person-level table `M`. Let's start with a clean question: **does hyperlipidemia become more likely with age, and does sex matter?** A logistic regression (`family = "binomial"`), fitted jointly across the three hospitals:

```{r glm1, results='hide'}
fit1 <- ds.glm(
  formula = "M$hyperlipidemia ~ M$age + M$sex",
  family  = "binomial", datasources = conns)
```

```{r glm1-table}
or_table <- function(fit) {
  co <- fit$coefficients
  data.frame(
    term    = rownames(co),
    OR      = round(co[, "P_OR"], 3),
    CI_low  = round(co[, "low0.95CI.P_OR"], 3),
    CI_high = round(co[, "high0.95CI.P_OR"], 3),
    p_value = signif(co[, "p-value"], 3),
    row.names = NULL
  )
}
or_table(fit1)
```

::: {.callout-important}
## Reading the result (`OR` = odds ratio)
- **age** is a clear, **significant** predictor — the odds of hyperlipidemia rise a few percent per year of age (OR > 1, p well below 0.05, confidence interval entirely above 1).
- **sex** is **not** significant here (OR ≈ 1, p ≈ 1, CI straddling 1).

That is exactly how you read a `ds.glm` table: one clear effect, one null — produced without ever touching a patient record.
:::

Now a **multivariable** model: does hypertension depend on age and sex **after adjusting for** hyperlipidemia?

```{r glm2, results='hide'}
fit2 <- ds.glm(
  formula = "M$hypertension ~ M$age + M$sex + M$hyperlipidemia",
  family  = "binomial", datasources = conns)
```

```{r glm2-table}
or_table(fit2)
```

::: {.callout-important}
## Adjustment, and the limits of a small sample
Age stays a significant predictor of hypertension; sex is **borderline**; and the hyperlipidemia adjustment is not significant. On a **~100-patient** demo, power is limited — "estimable and converged" is the success criterion here, not a definitive clinical claim. The companion GiBleed study (next) has ~2,700 patients and delivers a strongly significant result.
:::

```{r logout}
DSI::datashield.logout(conns)
```

---

# Where to go next

You now have the full toolkit: **connect → explore (conditions, numeric shape, categorical levels) → recipe (variables / format / filters / output, single or multiple) → preview → execute → ratify → summarise → model**, all under a disclosure filter that kept every patient private.

::: {.callout-note}
## Coming next
The **[exercise sheet](omop-exercises.html)** lets you practise everything here on new variables and a new model (with hints and a worked solution). Then a second vignette takes these exact tools to the larger **GiBleed** dataset (~2,700 patients) — where, with no numeric values to lean on, everything is treated as **boolean presence** — for a more powerful study.
:::
