---
title: "Federated OMOP analysis with dsOMOP"
subtitle: "A hands-on tutorial — exploring real hospital data across three sites, without the data ever leaving them"
format:
  html:
    code-copy: true
execute:
  warning: false
  message: false
  error: true
---

<a href="#" onclick="event.preventDefault(); fetch('https://raw.githubusercontent.com/isglobal-brge/workshop_DSWB/main/dsomop.R').then(response => response.blob()).then(blob => { const url = window.URL.createObjectURL(blob); const a = document.createElement('a'); a.style.display = 'none'; a.href = url; a.download = 'dsomop.R'; document.body.appendChild(a); a.click(); window.URL.revokeObjectURL(url); }); return false;" class="btn btn-primary btn-sm">
<i class="bi bi-download"></i> Download the R script (.R)
</a>
<a href="#" onclick="event.preventDefault(); fetch('https://raw.githubusercontent.com/isglobal-brge/workshop_DSWB/main/dsomop.Rmd').then(response => response.blob()).then(blob => { const url = window.URL.createObjectURL(blob); const a = document.createElement('a'); a.style.display = 'none'; a.href = url; a.download = 'dsomop.Rmd'; document.body.appendChild(a); a.click(); window.URL.revokeObjectURL(url); }); return false;" class="btn btn-outline-primary btn-sm">
<i class="bi bi-download"></i> Download the source (.Rmd)
</a>

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(comment = "#>", collapse = FALSE)
# small helper: render a data.frame as a clean table
show_tbl <- function(x, ...) knitr::kable(x, ...)
```

# Welcome

In this tutorial you will analyse **real-world hospital data** that lives in **three different hospitals at once** — and you will do it **without ever seeing or downloading a single patient's record**.

The pieces:

- **OMOP CDM** is a standard *shape* for health data. Every hospital stores its patients, diagnoses, drugs and lab results in the same tables (`person`, `condition_occurrence`, `drug_exposure`, `measurement`, ...). Learn it once, use it anywhere.
- **DataSHIELD** is the safety layer. You send *commands* to each hospital; the hospital runs them on its own data and returns **only aggregated, non-disclosive results**. Individual rows never travel.
- **dsOMOP** is the bridge that lets DataSHIELD speak OMOP.

::: {.callout-important}
## The golden rule
You never hold patient-level data on your laptop. Every result is checked by a **disclosure filter** on each server before it is returned. If a result would describe too few people, the server hides it. This explains much of what you'll see.
:::

We work with **MIMIC** (an intensive-care database) throughout this tutorial. You will: connect → explore → learn the **recipe** for extracting analysis tables → filter at extraction → and finish with a real **regression model** — all federated across three sites.

---

# 1. Connecting to the federation

```{r libs}
library(DSI)
library(DSOpal)
library(dsBaseClient)
library(dsOMOPClient)
```

Describe the three servers and log in (same training login on each, asking for the **`omop`** profile), then attach the MIMIC database under the symbol `"omop"`:

```{r login}
builder <- DSI::newDSLoginBuilder()
builder$append(server = "nairobi", url = "https://nairobi.datashield.live",
               user = "ethiopia", password = "P@ssw0rd", profile = "omop")
builder$append(server = "douala",  url = "https://douala.datashield.live",
               user = "ethiopia", password = "P@ssw0rd", profile = "omop")
builder$append(server = "dakar",   url = "https://dakar.datashield.live",
               user = "ethiopia", password = "P@ssw0rd", profile = "omop")
conns <- DSI::datashield.login(logins = builder$build())

ds.omop.connect(resource = "omop_demo.mimic", symbol = "omop", conns = conns)
```

::: {.callout-warning}
Forgetting `profile = "omop"` is the #1 beginner mistake — without it the `ds.omop.*` functions aren't available. If that happens, log out and rebuild the login with the profile set.
:::

---

# 2. Reading the map: tables and columns

Ask what tables exist (the result has one entry per server; the schemas are identical, so we peek at one):

```{r tables}
tabs <- ds.omop.tables(symbol = "omop", conns = conns)
show_tbl(head(tabs$nairobi[, c("table_name", "schema_category", "has_person_id")], 12))
```

Inspect a table's columns:

```{r columns}
show_tbl(ds.omop.columns("measurement", symbol = "omop", conns = conns)$nairobi)
```

How big is a table? `ds.omop.table.stats()` returns safe row/patient counts:

```{r tablestats}
st <- ds.omop.table.stats("condition_occurrence", symbol = "omop", conns = conns)$per_site$nairobi
show_tbl(data.frame(rows = st$rows, persons = st$persons))
```

::: {.callout-tip collapse="true"}
## Where would you look for...?
diagnoses → `condition_occurrence`; medications → `drug_exposure`; lab values / vital signs → `measurement`; social facts → `observation`. Every one has a `person_id` linking a patient's records.
:::

---

# 3. Who are these patients?

Gender lives in `person.gender_concept_id`. Rank its values, pooled across all three sites:

```{r gender}
g <- ds.omop.concept.prevalence("person", concept_col = "gender_concept_id",
                                scope = "pooled", symbol = "omop", conns = conns)
show_tbl(g$pooled)
```

::: {.callout-important}
## Concepts translate themselves
You got `MALE` / `FEMALE`, not codes like `8507` / `8532`. Any `*_concept_id` column is automatically translated to readable labels — **concept translation is on by default** everywhere in dsOMOP. You never memorise codes.
:::

::: {.callout-note}
## A real-world aside: de-identification
Intensive-care data is de-identified by shifting dates. For this teaching dataset the dates have been re-anchored so ages are realistic, but remember: in real clinical data, **dates are often deliberately distorted while ages and time-intervals are preserved**. Always derive age from the data, never assume the calendar year means "now". (We'll get proper age from the recipe in Section 7.)
:::

---

# 4. Discovering what was recorded

The most common diagnoses (ranked by distinct patients):

```{r topcond}
pc <- ds.omop.concept.prevalence("condition_occurrence", metric = "persons",
                                 top_n = 10, scope = "pooled", symbol = "omop", conns = conns)
show_tbl(pc$pooled)
```

Rows with `suppressed = TRUE` describe too few patients to report safely — the disclosure filter at work.

Don't know a concept's code? **Search** for it:

```{r search}
show_tbl(ds.omop.concept.search("hypertension", domain = "Condition",
                                limit = 5, symbol = "omop", conns = conns))
```

And **look up** codes you already have:

```{r lookup}
show_tbl(ds.omop.concept.lookup(c(320128, 432867, 3027018),
                                symbol = "omop", conns = conns))
```

::: {.callout-tip collapse="true"}
## Your turn
Find the `concept_id` for a condition that interests you with `ds.omop.concept.search(..., domain = "Condition")`.
:::

---

# 5. A numeric variable: heart rate

Measurements store a number in `value_as_number`. Take **heart rate** (`concept_id = 3027018`) and get a safe statistical summary:

```{r hr-stats}
cs <- ds.omop.column.stats("measurement", "value_as_number", concept_id = 3027018,
                           scope = "pooled", symbol = "omop", conns = conns)
show_tbl(cs$pooled)
```

::: {.callout-important}
You get `n`, `mean`, `sd` — but **not** the true min/max (exact extremes could identify a patient). When you need a range, use the disclosure-safe 5th–95th-percentile band.
:::

A picture of the distribution. `ds.omop.value.histogram()` returns disclosure-safe **bins** (per site); we combine the sites' bin counts and draw them:

```{r hr-hist, fig.width=7, fig.height=4}
h <- ds.omop.value.histogram("measurement", value_col = "value_as_number",
                             concept_id = 3027018, symbol = "omop", conns = conns)
bins <- do.call(rbind, h$per_site)            # stack the three sites' bins
bins <- subset(bins, !is.na(count))           # drop suppressed bins
bins <- aggregate(count ~ bin_start, data = bins, FUN = sum)
bins <- bins[order(bins$bin_start), ]
barplot(bins$count, names.arg = round(bins$bin_start), las = 2,
        col = "#4C72B0", border = NA,
        xlab = "Heart rate (bpm)", ylab = "records",
        main = "Heart rate across the federation")
```

::: {.callout-tip collapse="true"}
## What shape is it?
Roughly symmetric and unimodal, centred near the mean. Compare that with a skewed lab like serum creatinine (`3016723`), where a long right tail pulls the mean above the median.
:::

---

# 6. A categorical variable: heart rhythm

Some measurements store a *category* in `value_as_concept_id`. Heart **rhythm** (`concept_id = 3022318`) is the classic example. Count how often each value occurs:

```{r rhythm}
rh <- ds.omop.value.counts("measurement", "value_as_concept_id",
                           concept_id = 3022318, symbol = "omop", conns = conns)
show_tbl(rh$pooled)
```

::: {.callout-important}
## Two kinds of measurement
A **numeric** measurement (heart rate *value*) → summarise with mean / sd / histogram. A **categorical** one (heart *rhythm*) → count each level. And again the coded values came back as readable labels (`Sinus rhythm`, `Atrial fibrillation`, ...).
:::

---

# 7. The recipe: building an analysis table

Single aggregates are great for exploring, but to **model** anything we need a table with **one row per patient** and **one column per variable**. In dsOMOP you describe that table with a **recipe** and the servers build it for you — still without sending any rows.

A recipe has four elements:

| Element | What it is |
|---|---|
| **variables** | the columns you want — each an `omop_variable(...)` (a `person` column, or something built from a `concept_id`) |
| **filters** | optional row/population restrictions applied **during** extraction (Section 8) |
| **format** | *how* each variable is summarised per patient: `"sex_mf"`, `"binary"` (0/1 presence), `"count"`, `"mean"`, `"sd"`, ... (and `omop_variable_age()` builds age) |
| **output** | `omop_output(name, type = "wide")` — `"wide"` means one row per patient, exactly what a model needs |

Let's build a person-level table mixing all the variable types we've met — a categorical (`sex`), two numerics (`age`, mean `heart_rate`) and two binary condition flags (`hypertension` = 320128, `hyperlipidemia` = 432867):

```{r recipe-build}
rec <- omop_recipe(
  variables = list(
    omop_variable(table = "person", column = "gender_concept_id", format = "sex_mf", name = "sex"),
    omop_variable_age(name = "age"),
    omop_variable(table = "measurement", concept_id = 3027018, format = "mean",   name = "heart_rate"),
    omop_variable(table = "condition_occurrence", concept_id = 320128, format = "binary", name = "hypertension"),
    omop_variable(table = "condition_occurrence", concept_id = 432867, format = "binary", name = "hyperlipidemia")
  ),
  outputs = omop_output(name = "study", type = "wide")
)
```

**First, preview** — a server-side dry run that reports the expected columns and a *banded* patient count, **without building anything**:

```{r preview}
recipe_preview(rec, symbol = "omop", conns = conns)
```

**Then execute** — this builds the table on each server. We map the output named `study` to a server-side symbol `M`:

```{r execute}
recipe_execute(rec, out = c(study = "M"), symbol = "omop", conns = conns)
```

**Then check what you downloaded** — never trust a table you haven't inspected:

```{r check-cols}
ds.colnames("M", datasources = conns)
ds.dim("M", datasources = conns)
```

```{r check-summaries}
# numeric: a summary; categorical/binary: frequency tables
ds.summary("M$age", datasources = conns)[[1]][["quantiles & mean"]]
```

```{r check-tables}
tab_sex <- ds.table("M$sex", datasources = conns)$output.list[["TABLES.COMBINED_all.sources_counts"]]
show_tbl(as.data.frame(tab_sex))
tab_htn <- ds.table("M$hypertension", datasources = conns)$output.list[["TABLES.COMBINED_all.sources_counts"]]
show_tbl(as.data.frame(tab_htn))
```

---

# 8. Filtering at extraction

Often you only want a **subpopulation**. The cleanest way is to filter **while extracting** — add a `filters` element to the recipe so the servers only ever build the rows you asked for. Here we keep only patients **aged 65+**:

```{r recipe-filtered}
rec_old <- omop_recipe(
  variables = list(
    omop_variable(table = "person", column = "gender_concept_id", format = "sex_mf", name = "sex"),
    omop_variable_age(name = "age"),
    omop_variable(table = "condition_occurrence", concept_id = 320128, format = "binary", name = "hypertension")
  ),
  filters  = list(elderly = omop_filter_age(min = 65)),
  outputs  = omop_output(name = "study", type = "wide")
)
recipe_execute(rec_old, out = c(study = "E"), symbol = "omop", conns = conns)
```

**Check the filter did what you meant** — the minimum age should now be 65, and the cohort smaller:

```{r filtered-check}
ds.dim("E", datasources = conns)
ds.summary("E$age", datasources = conns)[[1]][["quantiles & mean"]]
```

::: {.callout-note}
Allowed filters are the *safe* ones — `omop_filter_age()`, ranges (`>=`, `>`, `<=`, `<`, `between`), and membership — because they can only ever **shrink** a population. The per-patient gate still runs on the result, so a filter can never bypass suppression.
:::

---

# 9. A real model: what predicts hypertension?

We have a person-level table `M`. Let's ask a genuine question: **does hypertension become more likely with age, and does sex or having hyperlipidemia matter?** A logistic regression (`family = "binomial"`), fitted jointly across the three hospitals:

```{r glm, results='hide'}
fit <- ds.glm(
  formula = "M$hypertension ~ M$age + M$sex + M$hyperlipidemia",
  family  = "binomial", datasources = conns)
```

```{r glm-table}
co <- fit$coefficients
res <- data.frame(
  term    = rownames(co),
  OR      = round(co[, "P_OR"], 2),
  CI_low  = round(co[, "low0.95CI.P_OR"], 2),
  CI_high = round(co[, "high0.95CI.P_OR"], 2),
  p_value = signif(co[, "p-value"], 3),
  row.names = NULL)
show_tbl(res)
```

::: {.callout-important}
## Reading the result (the `OR` column = odds ratio)
- **age** is a clear, significant predictor (odds of hypertension rise a few percent per year of age) — clinically sensible.
- **sex** and **hyperlipidemia** are *not* significant here (p well above 0.05, confidence intervals crossing 1).

A real study gives a **mixed** answer like this — one clear effect and some nulls. That's honest science, produced without ever touching a patient record. (On a ~100-patient demo, power is limited; the message is the *method*, not a definitive clinical claim.)
:::

```{r logout}
DSI::datashield.logout(conns)
```

---

# Where to go next

You now have the full toolkit: **connect → explore → summarise (numeric & categorical) → recipe (variables / filters / format / output) → preview → execute → check → model**, all under a disclosure filter that kept every patient private.

::: {.callout-note}
## Coming next
A second vignette takes these exact tools to a **larger dataset (GiBleed, ~2,700 patients)** to run a more powerful study — and the companion **exercise sheet** lets you practise everything here on new variables and a new model.
:::
