---
title: "GiBleed — Exercises"
subtitle: "Find your own significant association in a larger, boolean-only cohort"
format:
  html:
    df-print: kable
execute:
  eval: false
---

<a href="#" onclick="event.preventDefault(); fetch('https://raw.githubusercontent.com/isglobal-brge/workshop_DSWB/main/gibleed-exercises.R').then(response => response.blob()).then(blob => { const url = window.URL.createObjectURL(blob); const a = document.createElement('a'); a.style.display = 'none'; a.href = url; a.download = 'gibleed-exercises.R'; document.body.appendChild(a); a.click(); window.URL.revokeObjectURL(url); }); return false;" class="btn btn-primary btn-sm">
<i class="bi bi-download"></i> Download the worksheet (.R)
</a>
<a href="#" onclick="event.preventDefault(); fetch('https://raw.githubusercontent.com/isglobal-brge/workshop_DSWB/main/gibleed-exercises.Rmd').then(response => response.blob()).then(blob => { const url = window.URL.createObjectURL(blob); const a = document.createElement('a'); a.style.display = 'none'; a.href = url; a.download = 'gibleed-exercises.Rmd'; document.body.appendChild(a); a.click(); window.URL.revokeObjectURL(url); }); return false;" class="btn btn-outline-primary btn-sm">
<i class="bi bi-download"></i> Source (.Rmd)
</a>
<a href="gibleed_solution.html" target="_blank" class="btn btn-success btn-sm">
<i class="bi bi-check-circle"></i> View solutions
</a>

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = "#>", eval = FALSE)
```

# How to use this sheet

You did the whole workflow in the [GiBleed tutorial](gibleed.html). Now find a **different** association yourself. Fill the `____` blanks, run each chunk, and read the output. Remember: GiBleed has **no numeric values**, so every clinical variable is a **boolean** flag (`format = "binary"`) and age is anchored to **2019**. Worked, executed [solutions](gibleed_solution.html) are one click away.

## Setup (run this first)

```{r setup-conn}
library(DSI); library(DSOpal); library(dsBaseClient); library(dsOMOPClient)

builder <- DSI::newDSLoginBuilder()
builder$append(server = "nairobi", url = "https://nairobi.datashield.live",
               user = "ethiopia", password = "P@ssw0rd", profile = "omop")
builder$append(server = "douala",  url = "https://douala.datashield.live",
               user = "ethiopia", password = "P@ssw0rd", profile = "omop")
builder$append(server = "dakar",   url = "https://dakar.datashield.live",
               user = "ethiopia", password = "P@ssw0rd", profile = "omop")
conns <- DSI::datashield.login(logins = builder$build())
ds.omop.connect(resource = "omop_demo.gibleed", symbol = "omop", conns = conns)
```

---

# Exercise 1 — Explore the medications

The tutorial looked at *conditions*. Now explore **drugs**: which are most common, and how many patients take **celecoxib** (an anti-inflammatory / NSAID)?

```{r ge1}
# STEP 1 — pooled prevalence of the top drugs (by distinct patients)
ds.omop.concept.prevalence("____", metric = "persons", top_n = 12,
                           scope = "pooled", symbol = "omop", conns = conns)$pooled

# STEP 2 — confirm celecoxib's concept_id and how many patients take it
ds.omop.concept.lookup(1118084, symbol = "omop", conns = conns)$pooled
```

**Question:** Roughly how many patients are on celecoxib?

::: {.callout-tip collapse="true"}
## Hint
Drugs live in the **`drug_exposure`** table. Celecoxib is `concept_id = 1118084`.
:::

# Exercise 2 — Build a boolean table

Build a wide table `M` with **sex**, **age** (anchored to 2019), a boolean **GI-bleed** flag (`192671`) and a boolean **celecoxib** flag (`1118084`). Ratify it, and read each flag's prevalence.

```{r ge2}
rec <- omop_recipe(
  variables = list(
    omop_variable(table = "person", column = "gender_concept_id", format = "sex_mf", name = "sex"),
    omop_variable_age(name = "age", year = ____),
    omop_variable(table = "condition_occurrence", concept_id = 192671, format = "____", name = "gi_bleed"),
    omop_variable(table = "drug_exposure", concept_id = 1118084, format = "____", name = "celecoxib")
  ),
  output = omop_output(name = "study", type = "wide"))

recipe_execute(rec, out = c(study = "M"), symbol = "omop", conns = conns)
ds.colnames("M", datasources = conns); ds.dim("M", datasources = conns)
as.data.frame(ds.table("M$celecoxib", datasources = conns)$output.list[["TABLES.COMBINED_all.sources_counts"]])
```

::: {.callout-tip collapse="true"}
## Hint
`year = 2019`; both clinical flags are `format = "binary"` (a yes/no presence).
:::

# Exercise 3 — Is celecoxib associated with GI bleeding?

Fit a logistic regression of GI bleed on age, sex and celecoxib.

```{r ge3}
fit <- ds.glm(formula = "M$gi_bleed ~ M$age + M$sex + M$____",
              family = "binomial", datasources = conns)
fit$coefficients
```

**Question:** Is celecoxib a significant predictor (p < 0.05)? In which direction (OR above or below 1)?

::: {.callout-tip collapse="true"}
## Hint
The third predictor is `M$celecoxib`. Expect a **significant** odds ratio **above 1** (an NSAID raising GI-bleed odds). Age and sex should be non-significant in this older cohort.
:::

# Exercise 4 — A different outcome: does age predict a fracture?

Unlike GI bleed, some outcomes *do* rise with age. Build a table with sex, age, and a boolean **forearm-fracture** flag (`concept_id = 4278672`), then model it.

```{r ge4}
rec_fx <- omop_recipe(
  variables = list(
    omop_variable(table = "person", column = "gender_concept_id", format = "sex_mf", name = "sex"),
    omop_variable_age(name = "age", year = 2019),
    omop_variable(table = "condition_occurrence", concept_id = 4278672, format = "binary", name = "fx_forearm")
  ),
  output = omop_output(name = "study", type = "wide"))
recipe_execute(rec_fx, out = c(study = "FX"), symbol = "omop", conns = conns)

fit_fx <- ds.glm(formula = "FX$fx_forearm ~ FX$____ + FX$____",
                 family = "binomial", datasources = conns)
fit_fx$coefficients
```

**Question:** Which predictor is significant now? How does that contrast with the GI-bleed model?

::: {.callout-tip collapse="true"}
## Hint
Predictors are `FX$age` and `FX$sex`. Here **age** should be significant — older patients fall more — whereas in the GI-bleed model the *comorbidity*, not age, carried the signal.
:::

# Open challenge

Pick any condition or drug from the tutorial's prevalence chart, add it as a boolean flag, and test whether it predicts GI bleed. **First check its prevalence** (Section 7 of the tutorial showed what happens to a near-100% or near-0% outcome).

```{r challenge}
# YOUR CODE HERE
```

```{r logout}
DSI::datashield.logout(conns)
```

::: {.callout-note}
A larger cohort means more power — but also remember the per-patient disclosure gate still runs on every result, and a degenerate (all-yes / all-no) outcome still can't be modelled.
:::
