---
title: "GiBleed — Exercises (Solutions)"
subtitle: "Worked answers with real federated output"
format:
  html:
    df-print: kable
execute:
  warning: false
  message: false
  error: true
---

<a href="#" onclick="event.preventDefault(); fetch('https://raw.githubusercontent.com/isglobal-brge/workshop_DSWB/main/gibleed_solution.R').then(response => response.blob()).then(blob => { const url = window.URL.createObjectURL(blob); const a = document.createElement('a'); a.style.display = 'none'; a.href = url; a.download = 'gibleed_solution.R'; document.body.appendChild(a); a.click(); window.URL.revokeObjectURL(url); }); return false;" class="btn btn-primary btn-sm">
<i class="bi bi-download"></i> Download the solutions (.R)
</a>
<a href="gibleed-exercises.html" class="btn btn-outline-secondary btn-sm">
<i class="bi bi-arrow-left"></i> Back to the exercises
</a>

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = "#>")
```

```{r connect}
library(DSI); library(DSOpal); library(dsBaseClient); library(dsOMOPClient)

builder <- DSI::newDSLoginBuilder()
builder$append(server = "nairobi", url = "https://nairobi.datashield.live",
               user = "ethiopia", password = "P@ssw0rd", profile = "omop")
builder$append(server = "douala",  url = "https://douala.datashield.live",
               user = "ethiopia", password = "P@ssw0rd", profile = "omop")
builder$append(server = "dakar",   url = "https://dakar.datashield.live",
               user = "ethiopia", password = "P@ssw0rd", profile = "omop")
conns <- DSI::datashield.login(logins = builder$build())
ds.omop.connect(resource = "omop_demo.gibleed", symbol = "omop", conns = conns)

or_table <- function(fit) {
  co <- fit$coefficients
  data.frame(term = rownames(co),
             OR = round(co[, "P_OR"], 3),
             CI_low = round(co[, "low0.95CI.P_OR"], 3),
             CI_high = round(co[, "high0.95CI.P_OR"], 3),
             p_value = signif(co[, "p-value"], 3),
             row.names = NULL)
}
```

# Exercise 1 — Explore the medications

```{r ge1}
ds.omop.concept.prevalence("drug_exposure", metric = "persons", top_n = 12,
                           scope = "pooled", symbol = "omop", conns = conns)$pooled

ds.omop.concept.lookup(1118084, symbol = "omop", conns = conns)$pooled
```

> **Answer.** Celecoxib (`1118084`) is taken by **~1,840** patients — one of the most common drugs in this NSAID-heavy synthetic cohort.

# Exercise 2 — Build a boolean table

```{r ge2}
rec <- omop_recipe(
  variables = list(
    omop_variable(table = "person", column = "gender_concept_id", format = "sex_mf", name = "sex"),
    omop_variable_age(name = "age", year = 2019),
    omop_variable(table = "condition_occurrence", concept_id = 192671, format = "binary", name = "gi_bleed"),
    omop_variable(table = "drug_exposure", concept_id = 1118084, format = "binary", name = "celecoxib")
  ),
  output = omop_output(name = "study", type = "wide"))

recipe_execute(rec, out = c(study = "M"), symbol = "omop", conns = conns)
ds.colnames("M", datasources = conns)
ds.dim("M", datasources = conns)
as.data.frame(ds.table("M$celecoxib", datasources = conns)$output.list[["TABLES.COMBINED_all.sources_counts"]])
```

> **Answer.** `M` has one row per patient (~2,700 combined) with `person_id, sex, age, gi_bleed, celecoxib`. Both clinical columns are `binary` (0/1 presence); the server built exactly the table the recipe described.

# Exercise 3 — Is celecoxib associated with GI bleeding?

```{r ge3, results='hide'}
fit <- ds.glm(formula = "M$gi_bleed ~ M$age + M$sex + M$celecoxib",
              family = "binomial", datasources = conns)
```

```{r ge3-table}
or_table(fit)
```

> **Answer.** **Celecoxib is a significant predictor** — its odds ratio is **above 1** (≈ 1.4, p ≈ 0.003): patients on this NSAID have higher odds of a GI bleed, which is clinically sensible. Age and sex are not significant. (This is a *different* signal from the tutorial's peptic-ulcer model — a drug exposure rather than a prior condition.)

# Exercise 4 — A different outcome: does age predict a fracture?

```{r ge4, results='hide'}
rec_fx <- omop_recipe(
  variables = list(
    omop_variable(table = "person", column = "gender_concept_id", format = "sex_mf", name = "sex"),
    omop_variable_age(name = "age", year = 2019),
    omop_variable(table = "condition_occurrence", concept_id = 4278672, format = "binary", name = "fx_forearm")
  ),
  output = omop_output(name = "study", type = "wide"))
recipe_execute(rec_fx, out = c(study = "FX"), symbol = "omop", conns = conns)
fit_fx <- ds.glm(formula = "FX$fx_forearm ~ FX$age + FX$sex", family = "binomial", datasources = conns)
```

```{r ge4-table}
or_table(fit_fx)
```

> **Answer.** Now **age** is the significant predictor (OR slightly above 1 per year — older patients fracture more), while sex is not. The contrast with Exercise 3 is the lesson: *which* predictor matters depends on the outcome — a comorbidity drove GI bleeding, age drives fractures.

```{r logout}
DSI::datashield.logout(conns)
```
