---
title: "GiBleed — Exercises (Solutions)"
subtitle: "One worked path through each exercise, with real federated output"
format: html
execute:
  warning: false
  message: false
  error: true
---

<a href="#" onclick="event.preventDefault(); fetch('https://raw.githubusercontent.com/isglobal-brge/workshop_DSWB/main/gibleed_solution.R').then(response => response.blob()).then(blob => { const url = window.URL.createObjectURL(blob); const a = document.createElement('a'); a.style.display = 'none'; a.href = url; a.download = 'gibleed_solution.R'; document.body.appendChild(a); a.click(); window.URL.revokeObjectURL(url); }); return false;" class="btn btn-primary btn-sm">
<i class="bi bi-download"></i> Download the solutions (.R)
</a>
<a href="gibleed-exercises.html" class="btn btn-outline-secondary btn-sm">
<i class="bi bi-arrow-left"></i> Back to the exercises
</a>

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = "#>")
options(width = 200)   # print wide output on one line; the page scrolls it horizontally
```

```{r connect}
library(DSI); library(DSOpal); library(dsBaseClient); library(dsOMOPClient)

builder <- DSI::newDSLoginBuilder()
builder$append(server = "nairobi", url = "https://nairobi.datashield.live",
               user = "ethiopia", password = "P@ssw0rd", profile = "omop")
builder$append(server = "douala",  url = "https://douala.datashield.live",
               user = "ethiopia", password = "P@ssw0rd", profile = "omop")
builder$append(server = "dakar",   url = "https://dakar.datashield.live",
               user = "ethiopia", password = "P@ssw0rd", profile = "omop")
conns <- DSI::datashield.login(logins = builder$build())
ds.omop.connect(resource = "omop_demo.gibleed", symbol = "omop", conns = conns)
```

These are *one* valid path; your concepts/variables may differ. (GiBleed has no numeric values, so everything is a boolean flag; age is anchored to 2019.)

# Exercise 1 — Find a medication

```{r ex1}
ds.omop.concept.search("celecoxib", domain = "Drug",
                       symbol = "omop", conns = conns)

ds.omop.concept.prevalence("drug_exposure", metric = "persons", top_n = 12,
                           scope = "pooled", symbol = "omop", conns = conns)
```

> Celecoxib (an NSAID) is concept **1118084**, taken by ~1,840 patients.

# Exercise 2 — Does that drug relate to GI bleeding?

```{r ex2}
ds.omop.concept.search("gastrointestinal hemorrhage", domain = "Condition",
                       symbol = "omop", conns = conns)

rec <- omop_recipe(
  variables = list(
    omop_variable(table = "person", column = "gender_concept_id", format = "sex_mf", name = "sex"),
    omop_variable_age(name = "age", year = 2019),
    omop_variable(table = "condition_occurrence", concept_id = 192671, format = "binary", name = "gi_bleed"),
    omop_variable(table = "drug_exposure", concept_id = 1118084, format = "binary", name = "celecoxib")
  ),
  output = omop_output(name = "study", type = "wide"))
recipe_execute(rec, out = c(study = "M"), symbol = "omop", conns = conns)

fit <- ds.glm(formula = "M$gi_bleed ~ M$age + M$sex + M$celecoxib",
              family = "binomial", datasources = conns)
fit$coefficients
```

> GI haemorrhage is concept **192671**. Celecoxib is a **significant** predictor: its `P_OR` is above 1 (~1.4, `p` ≈ 0.003) — an NSAID raising GI-bleed odds. Age and sex are not significant.

# Exercise 3 — Does age predict a fracture?

```{r ex3}
ds.omop.concept.search("fracture of forearm", domain = "Condition",
                       symbol = "omop", conns = conns)

rec_fx <- omop_recipe(
  variables = list(
    omop_variable(table = "person", column = "gender_concept_id", format = "sex_mf", name = "sex"),
    omop_variable_age(name = "age", year = 2019),
    omop_variable(table = "condition_occurrence", concept_id = 4278672, format = "binary", name = "fx_forearm")
  ),
  output = omop_output(name = "study", type = "wide"))
recipe_execute(rec_fx, out = c(study = "FX"), symbol = "omop", conns = conns)

fit_fx <- ds.glm(formula = "FX$fx_forearm ~ FX$age + FX$sex", family = "binomial", datasources = conns)
fit_fx$coefficients
```

> Forearm fracture is concept **4278672**. Now **age** is the significant predictor (`P_OR` slightly above 1 — older patients fracture more), while sex is not. The contrast with Exercise 2 is the point: which predictor matters depends on the outcome.

```{r logout}
DSI::datashield.logout(conns)
```
