---
title: "GiBleed — Exercises"
subtitle: "Find your own significant association in a larger, boolean-only cohort"
format: html
execute:
  eval: false
---

<a href="#" onclick="event.preventDefault(); fetch('https://raw.githubusercontent.com/isglobal-brge/workshop_DSWB/main/gibleed-exercises.R').then(response => response.blob()).then(blob => { const url = window.URL.createObjectURL(blob); const a = document.createElement('a'); a.style.display = 'none'; a.href = url; a.download = 'gibleed-exercises.R'; document.body.appendChild(a); a.click(); window.URL.revokeObjectURL(url); }); return false;" class="btn btn-primary btn-sm">
<i class="bi bi-download"></i> Download the worksheet (.R)
</a>
<a href="gibleed_solution.html" target="_blank" class="btn btn-success btn-sm">
<i class="bi bi-check-circle"></i> View solutions
</a>

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = "#>", eval = FALSE)
```

# How to use this sheet

You did the workflow in the [GiBleed tutorial](gibleed.html). Now find a **different** association yourself, **looking up the concepts** with `ds.omop.concept.search(...)` (we won't hand you the codes). Each chunk is a bare skeleton plus a `# STEP` comment. Remember: GiBleed has **no numeric values**, so every clinical variable is a **boolean** flag (`format = "binary"`), and age is anchored to **2019**.

## Setup (run this first)

```{r setup-conn}
library(DSI); library(DSOpal); library(dsBaseClient); library(dsOMOPClient)

builder <- DSI::newDSLoginBuilder()
builder$append(server = "nairobi", url = "https://nairobi.datashield.live",
               user = "ethiopia", password = "P@ssw0rd", profile = "omop")
builder$append(server = "douala",  url = "https://douala.datashield.live",
               user = "ethiopia", password = "P@ssw0rd", profile = "omop")
builder$append(server = "dakar",   url = "https://dakar.datashield.live",
               user = "ethiopia", password = "P@ssw0rd", profile = "omop")
conns <- DSI::datashield.login(logins = builder$build())
ds.omop.connect(resource = "omop_demo.gibleed", symbol = "omop", conns = conns)
```

---

# Exercise 1 — Find a medication

Find a common **anti-inflammatory (NSAID) painkiller** in this cohort and how many patients take it.

```{r ex1}
# STEP 1 — search the Drug domain for an NSAID (e.g. one whose name you know)
ds.omop.concept.search()

# STEP 2 — how many patients are on it?
ds.omop.concept.prevalence()
```

# Exercise 2 — Does that drug relate to GI bleeding?

Find the concept for **gastrointestinal haemorrhage**, build a boolean table with sex, age, the GI-bleed flag and the drug flag, then model it.

```{r ex2}
# STEP 1 — find the GI-haemorrhage condition concept_id
ds.omop.concept.search()

# STEP 2 — wide table of booleans (choose the right formats + the 2019 age anchor)
omop_recipe(
  variables = list(
    omop_variable(),
    omop_variable_age(),
    omop_variable(),
    omop_variable()
  ),
  output = omop_output()
)
recipe_execute()

# STEP 3 — logistic regression; is the drug a significant predictor? (P_OR in $coefficients)
ds.glm()
```

# Exercise 3 — A different outcome: does age predict a fracture?

Unlike GI bleed, some outcomes *do* rise with age. Find a **forearm-fracture** condition and test whether age predicts it.

```{r ex3}
# STEP 1 — find the forearm-fracture concept_id
ds.omop.concept.search()

# STEP 2 — table with sex + age + the fracture flag, then a logistic model on age + sex
omop_recipe(
  variables = list(
    omop_variable(),
    omop_variable_age(),
    omop_variable()
  ),
  output = omop_output()
)
recipe_execute()
ds.glm()
```

```{r logout}
DSI::datashield.logout(conns)
```

::: {.callout-note}
Before modelling any outcome, check its prevalence (`ds.omop.concept.prevalence`) — a variable that everyone (or no one) has can't be modelled.
:::
