---
title: "Federated analysis on GiBleed"
subtitle: "A larger, boolean-only OMOP dataset — does a prior peptic ulcer raise the odds of GI bleeding?"
format: html
execute:
  warning: false
  message: false
  error: true
---

<a href="#" onclick="event.preventDefault(); fetch('https://raw.githubusercontent.com/isglobal-brge/workshop_DSWB/main/gibleed.R').then(response => response.blob()).then(blob => { const url = window.URL.createObjectURL(blob); const a = document.createElement('a'); a.style.display = 'none'; a.href = url; a.download = 'gibleed.R'; document.body.appendChild(a); a.click(); window.URL.revokeObjectURL(url); }); return false;" class="btn btn-primary btn-sm">
<i class="bi bi-download"></i> Download the R script (.R)
</a>
<a href="#" onclick="event.preventDefault(); fetch('https://raw.githubusercontent.com/isglobal-brge/workshop_DSWB/main/gibleed.Rmd').then(response => response.blob()).then(blob => { const url = window.URL.createObjectURL(blob); const a = document.createElement('a'); a.style.display = 'none'; a.href = url; a.download = 'gibleed.Rmd'; document.body.appendChild(a); a.click(); window.URL.revokeObjectURL(url); }); return false;" class="btn btn-outline-primary btn-sm">
<i class="bi bi-download"></i> Source (.Rmd)
</a>

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = "#>", collapse = FALSE)
options(width = 200)   # print wide output on one line; the page scrolls it horizontally
```

# Welcome

In the [first tutorial](dsomop.html) you learned the whole dsOMOP workflow on **MIMIC** (~100 ICU patients). Here we take **exactly the same tools** to a second, **larger** dataset — **GiBleed** (~2,700 patients) — for a more powerful study, and reproduce a classic federated finding:

> **Does a prior history of peptic ulcer raise the odds of a gastrointestinal (GI) bleed?**

::: {.callout-important}
## What's different about GiBleed
GiBleed is a synthetic OHDSI cohort with **no numeric lab/vital values** — `measurement.value_as_number` is essentially empty. So there is nothing to take a mean or histogram of. Instead, every clinical variable becomes a **boolean** ("does this patient *have* this condition / drug, yes or no?") or a **count** ("how many records?"). Same recipe machinery, different `format`.
:::

# 1. Connect

```{r libs}
library(DSI); library(DSOpal); library(dsBaseClient); library(dsOMOPClient)
```

```{r login}
builder <- DSI::newDSLoginBuilder()
builder$append(server = "nairobi", url = "https://nairobi.datashield.live",
               user = "ethiopia", password = "P@ssw0rd", profile = "omop")
builder$append(server = "douala",  url = "https://douala.datashield.live",
               user = "ethiopia", password = "P@ssw0rd", profile = "omop")
builder$append(server = "dakar",   url = "https://dakar.datashield.live",
               user = "ethiopia", password = "P@ssw0rd", profile = "omop")
conns <- DSI::datashield.login(logins = builder$build())

ds.omop.connect(resource = "omop_demo.gibleed", symbol = "omop", conns = conns)
```

Note the resource is now `omop_demo.gibleed` (not `mimic`).

# 2. Confirm there are no numeric values

Before choosing how to model, **check the data**. Ask for a numeric summary of any measurement — it comes back empty, which is *why* we will treat everything as boolean:

```{r no-numeric}
ds.omop.column.stats("measurement", "value_as_number",
                     scope = "pooled", symbol = "omop", conns = conns)
```

::: {.callout-note}
An `NA` mean / zero distinct values confirms `value_as_number` carries no data here. GiBleed records *that* an event happened, not a measured quantity — so our variables will be **presence flags** and **counts**, never means.
:::

# 3. The population

GiBleed is bigger than MIMIC. Sex, pooled across the three sites:

```{r sex}
ds.omop.concept.prevalence("person", concept_col = "gender_concept_id",
                           scope = "pooled", symbol = "omop", conns = conns)
```

::: {.callout-note}
## A reference year for age
GiBleed is a **frozen ~2019 Synthea snapshot**, so we anchor age to **2019** (`omop_variable_age(year = 2019)`) — the dataset's own era, not "today". (For MIMIC we used 2024; each dataset gets its own collection year.)
:::

# 4. Explore what was recorded

No numbers to plot — so our "distribution picture" is a **prevalence bar chart**: how many patients have each of the most common conditions.

```{r cond-prev, fig.width=8, fig.height=4.5}
cond <- ds.omop.concept.prevalence("condition_occurrence", metric = "persons",
                                   top_n = 12, scope = "pooled", symbol = "omop", conns = conns)$pooled
cond <- cond[order(cond$n_persons), ]
par(mar = c(4, 17, 2, 1))
barplot(cond$n_persons, names.arg = cond$concept_name, horiz = TRUE, las = 1,
        col = "#4C72B0", border = NA, xlab = "patients",
        main = "Most common conditions in GiBleed")
```

The two conditions for our question are in that list, but we don't have their codes yet — we only know their **names**. Turn a name into a concept_id with `concept.search`:

```{r find-concepts}
ds.omop.concept.search("gastrointestinal hemorrhage", domain = "Condition",
                       symbol = "omop", conns = conns)
ds.omop.concept.search("peptic ulcer", domain = "Condition",
                       symbol = "omop", conns = conns)
```

::: {.callout-tip}
## What we're after
- **GI hemorrhage** (`192671`) — our outcome.
- **Peptic ulcer** (`4027663`) — the antecedent condition we think raises the risk.

Both affect enough patients (hundreds each) to model. Notice the long list of near-universal Synthea complaints (osteoarthritis, sinusitis…) — we'll meet one of those again in Section 7.
:::

# 5. Build the analysis table (all boolean)

A wide person-level table with **sex**, **age** (anchored to 2019), and two **boolean** flags — GI bleed and peptic ulcer:

```{r recipe}
rec <- omop_recipe(
  variables = list(
    omop_variable(table = "person", column = "gender_concept_id", format = "sex_mf", name = "sex"),
    omop_variable_age(name = "age", year = 2019),
    omop_variable(table = "condition_occurrence", concept_id = 192671,  format = "binary", name = "gi_bleed"),
    omop_variable(table = "condition_occurrence", concept_id = 4027663, format = "binary", name = "peptic_ulcer")
  ),
  output = omop_output(name = "study", type = "wide")
)
recipe_execute(rec, out = c(study = "M"), symbol = "omop", conns = conns)
```

**Ratify what the server built**, then read off each flag's prevalence with a frequency table:

```{r check}
ds.colnames("M", datasources = conns)
ds.dim("M", datasources = conns)
ds.table("M$gi_bleed",     datasources = conns)
ds.table("M$peptic_ulcer", datasources = conns)
```

# 6. The model

Now the question, as a logistic regression across all three hospitals:

```{r glm, results='hide'}
fit <- ds.glm(
  formula = "M$gi_bleed ~ M$age + M$sex + M$peptic_ulcer",
  family  = "binomial", datasources = conns)
```

```{r glm-table}
fit$coefficients
```

::: {.callout-important}
## The result
A prior **peptic ulcer** multiplies the odds of GI bleeding roughly **four-fold** (OR ≈ 3.9, 95% CI well above 1, p astronomically small) — a strong, clinically coherent signal. **Age** and **sex** are not associated here. With ~2,700 patients the confidence interval is tight: this is the power a larger cohort buys you, and it was produced **without any patient record leaving its hospital**.
:::

::: {.callout-tip collapse="true"}
## Optional: reproduce the full published model
The reference study also adjusts for NSAID drugs. dsOMOP can build a **count across several drug concepts in one variable** — `omop_variable(table = "drug_exposure", concept_id = c(19059056, 1118084, 19078461, 1115171, 1124300), format = "count", name = "nsaid_n")` (aspirin, celecoxib, ibuprofen, naproxen, diclofenac) — and a single-drug flag, e.g. `diclofenac` (`1124300`, `format = "binary"`). Adding them (`gi_bleed ~ age + sex + peptic_ulcer + diclofenac + nsaid_n`) leaves **peptic ulcer dominant**; the drug terms are weak/noise in this synthetic cohort, so don't over-interpret them.
:::

# 7. A cautionary tale: when a model breaks

Not every variable can be modelled. **Osteoarthritis** (`80180`) is recorded for **every single patient** in GiBleed. Try to predict it:

```{r degenerate, warning=TRUE}
rec_oa <- omop_recipe(
  variables = list(
    omop_variable(table = "person", column = "gender_concept_id", format = "sex_mf", name = "sex"),
    omop_variable_age(name = "age", year = 2019),
    omop_variable(table = "condition_occurrence", concept_id = 80180, format = "binary", name = "osteoarthritis")
  ),
  output = omop_output(name = "study", type = "wide"))
recipe_execute(rec_oa, out = c(study = "OA"), symbol = "omop", conns = conns)

fit_oa <- ds.glm("OA$osteoarthritis ~ OA$age + OA$sex", family = "binomial", datasources = conns)
fit_oa$coefficients[, c("Estimate", "p-value")]
```

::: {.callout-important}
## Perfect separation
Because the outcome is **100% prevalent** (no "no" group to contrast against), the logistic regression **cannot converge** — you'll see a *"did not converge"* warning and a wildly inflated intercept. The lesson: **always check an outcome's prevalence before modelling it.** A variable everyone (or no one) has carries no information to fit.
:::

```{r logout}
DSI::datashield.logout(conns)
```

---

# Where to go next

You have now run the full federated workflow on **two** datasets — numeric-rich MIMIC and boolean-only GiBleed — and seen both a strong result and a broken one.

::: {.callout-note}
## Practise it
The **[GiBleed exercise sheet](gibleed-exercises.html)** asks you to find a *different* significant association in this cohort (with hints and a worked, executed [solution](gibleed_solution.html)).
:::
