---
title: "dsOMOP — Exercises"
subtitle: "Practise the federated-analysis toolkit on new MIMIC variables and a new model"
format:
  html:
    code-copy: true
execute:
  eval: false
---

<a href="#" onclick="event.preventDefault(); fetch('https://raw.githubusercontent.com/isglobal-brge/workshop_DSWB/main/omop-exercises.Rmd').then(response => response.blob()).then(blob => { const url = window.URL.createObjectURL(blob); const a = document.createElement('a'); a.style.display = 'none'; a.href = url; a.download = 'omop-exercises.Rmd'; document.body.appendChild(a); a.click(); window.URL.revokeObjectURL(url); }); return false;" class="btn btn-primary btn-sm">
<i class="bi bi-download"></i> Download this worksheet (.Rmd)
</a>
<a href="omop_exercises_solution.html" target="_blank" class="btn btn-success btn-sm">
<i class="bi bi-check-circle"></i> View solutions
</a>

# How to use this sheet

Everything you need you **already did once** in the tutorial. Each exercise has a blank code block (`# YOUR CODE HERE`) and a **collapsed hint** — try it yourself first. Write your answers to the questions in plain words. We stay on the **MIMIC** dataset throughout.

::: {.callout-important}
## Reminders
- Connect with `profile = "omop"`.
- Pooled summaries: pass `scope = "pooled"` and read the `$pooled` slot.
- **Check the output after every step.**
:::

## Setup (run this first)

```{r setup}
library(DSI); library(DSOpal); library(dsBaseClient); library(dsOMOPClient)

builder <- DSI::newDSLoginBuilder()
for (s in c("nairobi", "douala", "dakar"))
  builder$append(server = s, url = paste0("https://", s, ".datashield.live"),
                 user = "ethiopia", password = "P@ssw0rd", profile = "omop")
conns <- DSI::datashield.login(logins = builder$build())
ds.omop.connect(resource = "omop_demo.mimic", symbol = "omop", conns = conns)
```

---

# Exercise 1 — A new numeric variable

In the tutorial we summarised **heart rate**. Do the same for **systolic blood pressure** (`concept_id = 21492239`): get its safe statistics, then draw its histogram.

```{r e1}
# YOUR CODE HERE
```

**Question:** Is systolic blood pressure roughly symmetric, or skewed? How does its spread compare to heart rate's?

::: {.callout-tip collapse="true"}
## Hint
`ds.omop.column.stats("measurement", "value_as_number", concept_id = 21492239, scope = "pooled", symbol = "omop", conns = conns)$pooled` for the numbers. For the picture, reuse the tutorial's recipe: get the per-site bins with `ds.omop.value.histogram("measurement", value_col = "value_as_number", concept_id = 21492239, symbol = "omop", conns = conns)`, then aggregate `count` over `bin_start` and `barplot()`.
:::

# Exercise 2 — A new categorical variable

The tutorial used **heart rhythm**. Explore **marital status** (`concept_id = 40766231`, in the `observation` table): count how often each value occurs.

```{r e2}
# YOUR CODE HERE
```

**Question:** What is the most common marital status?

::: {.callout-tip collapse="true"}
## Hint
Marital status is a coded value, so it lives in `value_as_concept_id`:
`ds.omop.value.counts("observation", "value_as_concept_id", concept_id = 40766231, symbol = "omop", conns = conns)`
:::

# Exercise 3 — Find a concept yourself

You suspect **hyperlipidemia** is common. Find its `concept_id` by searching, then check how prevalent it is.

```{r e3}
# YOUR CODE HERE
```

**Question:** What is the concept_id, and how many patients have it?

::: {.callout-tip collapse="true"}
## Hint
`ds.omop.concept.search("hyperlipidemia", domain = "Condition", symbol = "omop", conns = conns)` to get the id, then look for it in `ds.omop.concept.prevalence("condition_occurrence", metric = "persons", top_n = 15, scope = "pooled", symbol = "omop", conns = conns)$pooled`.
:::

# Exercise 4 — Build your own recipe

Build a person-level table `D` with four columns: **sex**, **age**, the **mean systolic blood pressure**, and a **binary flag for hyperlipidemia** (the concept from Exercise 3). Preview it, execute it, then check the columns, dimensions and a summary.

```{r e4}
# YOUR CODE HERE
```

::: {.callout-tip collapse="true"}
## Hint
```r
rec <- omop_recipe(
  variables = list(
    omop_variable(table = "person", column = "gender_concept_id", format = "sex_mf", name = "sex"),
    omop_variable_age(name = "age"),
    omop_variable(table = "measurement", concept_id = 21492239, format = "mean", name = "systolic_bp"),
    omop_variable(table = "condition_occurrence", concept_id = 432867, format = "binary", name = "hyperlipidemia")),
  outputs = omop_output(name = "study", type = "wide"))
recipe_preview(rec, symbol = "omop", conns = conns)
recipe_execute(rec, out = c(study = "D"), symbol = "omop", conns = conns)
ds.colnames("D", datasources = conns); ds.dim("D", datasources = conns)
```
:::

# Exercise 5 — Filter at extraction

Re-extract, but this time keep only the **older patients (age ≥ 60)** as `D_old`, by adding a **filter** to the recipe. Then check the age range to confirm the filter worked.

```{r e5}
# YOUR CODE HERE
```

**Question:** How many patients remain, and what is the new minimum age?

::: {.callout-tip collapse="true"}
## Hint
Add `filters = list(older = omop_filter_age(min = 60))` to your `omop_recipe(...)` and execute into `D_old`, then `ds.dim("D_old", ...)` and `ds.summary("D_old$age", ...)[[1]]`. Filtering during extraction means the servers only ever build the rows you asked for.
:::

# Exercise 6 — A model

Using a person-level table with `sex`, `age` and `hyperlipidemia`, fit a logistic regression: **does the chance of hyperlipidemia change with age or differ by sex?**

```{r e6}
# YOUR CODE HERE
```

**Question:** Which predictors are significant (p < 0.05)? Is the result clinically sensible?

::: {.callout-tip collapse="true"}
## Hint
Build a recipe with `sex`, `omop_variable_age()` and a binary `hyperlipidemia` (432867), execute into `H`, then
`ds.glm("H$hyperlipidemia ~ H$age + H$sex", family = "binomial", datasources = conns)$coefficients`.
:::

# Open challenge

Pick **any** condition that interests you (use `ds.omop.concept.search`), add it to a recipe, and explore whether it is associated with age or sex. There is no single right answer — the skill is running the workflow correctly and reading the result honestly.

```{r challenge}
# YOUR CODE HERE
```

```{r logout}
DSI::datashield.logout(conns)
```

::: {.callout-note}
Not every hypothesis pans out — a non-significant result is still a real finding. Doing the analysis correctly matters more than getting stars next to your coefficients.
:::
