---
title: "dsOMOP — Exercises (MIMIC)"
subtitle: "Find the concepts yourself and build the analysis"
format: html
execute:
  eval: false
---

<a href="#" onclick="event.preventDefault(); fetch('https://raw.githubusercontent.com/isglobal-brge/workshop_DSWB/main/omop-exercises.R').then(response => response.blob()).then(blob => { const url = window.URL.createObjectURL(blob); const a = document.createElement('a'); a.style.display = 'none'; a.href = url; a.download = 'omop-exercises.R'; document.body.appendChild(a); a.click(); window.URL.revokeObjectURL(url); }); return false;" class="btn btn-primary btn-sm">
<i class="bi bi-download"></i> Download the worksheet (.R)
</a>
<a href="omop_exercises_solution.html" target="_blank" class="btn btn-success btn-sm">
<i class="bi bi-check-circle"></i> View solutions
</a>

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = "#>", eval = FALSE)
```

# How to use this sheet

You did the whole workflow in the [tutorial](dsomop.html). Now do it yourself on **new** variables — and **find the concept codes yourself** with `ds.omop.concept.search(...)`; we won't hand them to you. Each chunk is a **bare skeleton** (the functions you'll need, with empty arguments) plus a `# STEP` comment — fill it in and think about each argument. Worked, executed [solutions](omop_exercises_solution.html) are one click away. Anchor age with `year = 2024`.

## Setup (run this first)

```{r setup-conn}
library(DSI); library(DSOpal); library(dsBaseClient); library(dsOMOPClient)

builder <- DSI::newDSLoginBuilder()
builder$append(server = "nairobi", url = "https://nairobi.datashield.live",
               user = "ethiopia", password = "P@ssw0rd", profile = "omop")
builder$append(server = "douala",  url = "https://douala.datashield.live",
               user = "ethiopia", password = "P@ssw0rd", profile = "omop")
builder$append(server = "dakar",   url = "https://dakar.datashield.live",
               user = "ethiopia", password = "P@ssw0rd", profile = "omop")
conns <- DSI::datashield.login(logins = builder$build())
ds.omop.connect(resource = "omop_demo.mimic", symbol = "omop", conns = conns)
```

---

# Exercise 1 — Find a condition

How many patients have a recorded problem with **high blood cholesterol**? Find the concept, then its prevalence.

```{r ex1}
# STEP 1 — search the Condition domain for the right concept (what term describes high cholesterol?)
ds.omop.concept.search()

# STEP 2 — once you have its concept_id, see how many patients have it
ds.omop.concept.prevalence()
```

# Exercise 2 — A numeric measurement

Pick a **blood-pressure** measurement, get its disclosure-safe statistics, and plot its distribution. Is it symmetric or skewed?

```{r ex2}
# STEP 1 — find the concept_id for a systolic blood pressure measurement
ds.omop.concept.search()

# STEP 2 — safe summary (mean, sd, p05-p95)
ds.omop.column.stats()

# STEP 3 — draw it (remember the histogram can plot itself)
ds.omop.value.histogram()
```

# Exercise 3 — Are these patients married?

Find out the most common **marital situation** of the patients. (It's recorded in the `observation` table as a coded value.)

```{r ex3}
# STEP 1 — find the concept_id that records marital status
ds.omop.concept.search()

# STEP 2 — count each recorded value (which column holds a CODED answer?)
ds.omop.value.counts()
```

# Exercise 4 — Build your own table

Build a person-level table with **sex**, **age**, a **numeric** lab/vital of your choice, and a **yes/no flag** for a condition you found above. Then ratify it and summarise each column.

```{r ex4}
# STEP 1 — fill the recipe (choose formats: sex_mf / mean / binary / count ...)
omop_recipe(
  variables = list(
    omop_variable(),
    omop_variable_age(),
    omop_variable(),
    omop_variable()
  ),
  output = omop_output()
)

# STEP 2 — execute into a server symbol, then RATIFY (colnames + dim)
recipe_execute()

# STEP 3 — summarise: ds.summary() for numerics, ds.table() for categories/flags
```

# Exercise 5 — Filter, then model

Restrict to a subpopulation (e.g. older patients), then fit a logistic regression for a yes/no outcome of your choice. Read the raw `ds.glm` coefficients.

```{r ex5}
# STEP 1 — recipe with a filters = list(...) element, then execute
omop_recipe(
  variables = list(),
  filters = list(),
  output = omop_output()
)
recipe_execute()

# STEP 2 — fit the model and inspect $coefficients (P_OR is the odds ratio)
ds.glm()
```

```{r logout}
DSI::datashield.logout(conns)
```

::: {.callout-note}
Stuck on a concept? `ds.omop.concept.search("<term>", domain = "Condition")` (or `domain = "Measurement"`). A non-significant model is still a real result.
:::
