---
title: "dsOMOP — Exercises (Solutions)"
format:
  html:
    code-copy: true
execute:
  warning: false
  message: false
  error: true
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(comment = "#>")
show_tbl <- function(x, ...) knitr::kable(x, ...)
```

```{r connect}
library(DSI); library(DSOpal); library(dsBaseClient); library(dsOMOPClient)

builder <- DSI::newDSLoginBuilder()
builder$append(server = "nairobi", url = "https://nairobi.datashield.live",
               user = "ethiopia", password = "P@ssw0rd", profile = "omop")
builder$append(server = "douala",  url = "https://douala.datashield.live",
               user = "ethiopia", password = "P@ssw0rd", profile = "omop")
builder$append(server = "dakar",   url = "https://dakar.datashield.live",
               user = "ethiopia", password = "P@ssw0rd", profile = "omop")
conns <- DSI::datashield.login(logins = builder$build())
ds.omop.connect(resource = "omop_demo.mimic", symbol = "omop", conns = conns)
```

# Exercise 1 — A new numeric variable

```{r e1, fig.width=7, fig.height=4}
show_tbl(ds.omop.column.stats("measurement", "value_as_number", concept_id = 21492239,
                              scope = "pooled", symbol = "omop", conns = conns)$pooled)

h <- ds.omop.value.histogram("measurement", value_col = "value_as_number",
                             concept_id = 21492239, symbol = "omop", conns = conns)
bins <- do.call(rbind, h$per_site)
bins <- subset(bins, !is.na(count))
bins <- aggregate(count ~ bin_start, data = bins, FUN = sum)
bins <- bins[order(bins$bin_start), ]
barplot(bins$count, names.arg = round(bins$bin_start), las = 2, col = "#4C72B0", border = NA,
        xlab = "Systolic BP (mmHg)", ylab = "records", main = "Systolic blood pressure")
```

> **Answer.** Systolic blood pressure is roughly **symmetric** (mean ≈ median, ~115–120 mmHg), with a **wider spread** than heart rate (sd ≈ 22 vs ≈ 19). A broad-but-symmetric distribution is typical of blood pressure.

# Exercise 2 — A new categorical variable

```{r e2}
rh <- ds.omop.value.counts("observation", "value_as_concept_id", concept_id = 40766231,
                           symbol = "omop", conns = conns)
show_tbl(rh$per_site$nairobi)
```

> **Answer.** The most common status is **Never married**, followed by **Married** and **Widowed**. Some categories are suppressed on this ~100-patient cohort (small-cell control).

# Exercise 3 — Find a concept yourself

```{r e3}
show_tbl(ds.omop.concept.search("hyperlipidemia", domain = "Condition",
                                symbol = "omop", conns = conns))
```

> **Answer.** Hyperlipidemia is **`concept_id = 432867`**, present in about **45** of the ~100 patients — one of the most common conditions in this ICU cohort.

# Exercise 4 — Build your own recipe

```{r e4}
rec <- omop_recipe(
  variables = list(
    omop_variable(table = "person", column = "gender_concept_id", format = "sex_mf", name = "sex"),
    omop_variable_age(name = "age"),
    omop_variable(table = "measurement", concept_id = 21492239, format = "mean", name = "systolic_bp"),
    omop_variable(table = "condition_occurrence", concept_id = 432867, format = "binary", name = "hyperlipidemia")),
  outputs = omop_output(name = "study", type = "wide"))

recipe_execute(rec, out = c(study = "D"), symbol = "omop", conns = conns)
ds.colnames("D", datasources = conns)
ds.dim("D", datasources = conns)
ds.summary("D$systolic_bp", datasources = conns)[[1]][["quantiles & mean"]]
```

> **Answer.** `D` has one row per patient with `sex`, `age`, `systolic_bp`, `hyperlipidemia` (plus `person_id`); ~100 patients combined, mean systolic BP near 120 mmHg.

# Exercise 5 — Filter at extraction

```{r e5}
rec_old <- omop_recipe(
  variables = list(
    omop_variable(table = "person", column = "gender_concept_id", format = "sex_mf", name = "sex"),
    omop_variable_age(name = "age"),
    omop_variable(table = "condition_occurrence", concept_id = 432867, format = "binary", name = "hyperlipidemia")),
  filters  = list(older = omop_filter_age(min = 60)),
  outputs  = omop_output(name = "study", type = "wide"))
recipe_execute(rec_old, out = c(study = "D_old"), symbol = "omop", conns = conns)

ds.dim("D_old", datasources = conns)
ds.summary("D_old$age", datasources = conns)[[1]][["quantiles & mean"]]
```

> **Answer.** The filter is applied **during** extraction, so the servers only build patients aged 60+. The cohort shrinks and the minimum age is now 60.

# Exercise 6 — A model

```{r e6, results='hide'}
rec_h <- omop_recipe(
  variables = list(
    omop_variable(table = "person", column = "gender_concept_id", format = "sex_mf", name = "sex"),
    omop_variable_age(name = "age"),
    omop_variable(table = "condition_occurrence", concept_id = 432867, format = "binary", name = "hyperlipidemia")),
  outputs = omop_output(name = "study", type = "wide"))
recipe_execute(rec_h, out = c(study = "H"), symbol = "omop", conns = conns)
fit <- ds.glm("H$hyperlipidemia ~ H$age + H$sex", family = "binomial", datasources = conns)
```

```{r e6-table}
co <- fit$coefficients
show_tbl(data.frame(
  term = rownames(co), OR = round(co[, "P_OR"], 2),
  CI_low = round(co[, "low0.95CI.P_OR"], 2), CI_high = round(co[, "high0.95CI.P_OR"], 2),
  p_value = signif(co[, "p-value"], 3), row.names = NULL))
```

> **Answer.** **age** is a strong, significant predictor of hyperlipidemia (odds rise with age, p < 0.001) — clinically sensible, since cholesterol problems accumulate over a lifetime. **sex** is not significant. A clean, honest single-predictor result.

```{r logout}
DSI::datashield.logout(conns)
```
