---
title: "dsOMOP — Exercises (MIMIC)"
subtitle: "Practise the federated-analysis toolkit on new variables and a new model"
format:
  html:
    df-print: kable
execute:
  eval: false
---

<a href="#" onclick="event.preventDefault(); fetch('https://raw.githubusercontent.com/isglobal-brge/workshop_DSWB/main/omop-exercises.R').then(response => response.blob()).then(blob => { const url = window.URL.createObjectURL(blob); const a = document.createElement('a'); a.style.display = 'none'; a.href = url; a.download = 'omop-exercises.R'; document.body.appendChild(a); a.click(); window.URL.revokeObjectURL(url); }); return false;" class="btn btn-primary btn-sm">
<i class="bi bi-download"></i> Download the worksheet (.R)
</a>
<a href="#" onclick="event.preventDefault(); fetch('https://raw.githubusercontent.com/isglobal-brge/workshop_DSWB/main/omop-exercises.Rmd').then(response => response.blob()).then(blob => { const url = window.URL.createObjectURL(blob); const a = document.createElement('a'); a.style.display = 'none'; a.href = url; a.download = 'omop-exercises.Rmd'; document.body.appendChild(a); a.click(); window.URL.revokeObjectURL(url); }); return false;" class="btn btn-outline-primary btn-sm">
<i class="bi bi-download"></i> Source (.Rmd)
</a>
<a href="omop_exercises_solution.html" target="_blank" class="btn btn-success btn-sm">
<i class="bi bi-check-circle"></i> View solutions
</a>

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = "#>", eval = FALSE)
```

# How to use this sheet

Everything you need you **already did once** in the [tutorial](dsomop.html). Each exercise gives you a **code skeleton** with blanks marked `____` and a `# STEP` comment telling you what to do — fill in the blanks, run it, and **check the output**. A collapsed **hint** is there if you get stuck. We stay on the **MIMIC** dataset, but use *new* variables. The [worked solutions](omop_exercises_solution.html) (with real output) are one click away.

::: {.callout-important}
## Reminders
- Connect with `profile = "omop"`.
- Pooled summaries: pass `scope = "pooled"` and read the `$pooled` slot.
- Anchor age with `year = 2024` (this extract's collection year).
- **Check the output after every step.**
:::

## Setup (run this first)

```{r setup-conn}
library(DSI); library(DSOpal); library(dsBaseClient); library(dsOMOPClient)

builder <- DSI::newDSLoginBuilder()
builder$append(server = "nairobi", url = "https://nairobi.datashield.live",
               user = "ethiopia", password = "P@ssw0rd", profile = "omop")
builder$append(server = "douala",  url = "https://douala.datashield.live",
               user = "ethiopia", password = "P@ssw0rd", profile = "omop")
builder$append(server = "dakar",   url = "https://dakar.datashield.live",
               user = "ethiopia", password = "P@ssw0rd", profile = "omop")
conns <- DSI::datashield.login(logins = builder$build())
ds.omop.connect(resource = "omop_demo.mimic", symbol = "omop", conns = conns)
```

---

# Exercise 1 — Find a concept yourself

You suspect **alcohol abuse** is recorded. Find its `concept_id`, then see how many patients have it.

```{r e1}
# STEP 1 — search the Condition domain for "alcohol abuse" to get its concept_id
ds.omop.concept.search("____", domain = "Condition", symbol = "omop", conns = conns)$pooled

# STEP 2 — find it in the pooled condition prevalence (ranked by distinct patients)
ds.omop.concept.prevalence("condition_occurrence", metric = "persons",
                           top_n = 15, scope = "pooled", symbol = "omop", conns = conns)$pooled
```

**Question:** What is the `concept_id`, and roughly how many patients have it?

::: {.callout-tip collapse="true"}
## Hint
Search the term `"alcohol"`. You are looking for *Alcohol abuse* (`concept_id = 433753`); read its `n_persons` from the prevalence table.
:::

# Exercise 2 — A new numeric variable

The tutorial summarised heart rate and creatinine. Do the same for **systolic blood pressure** (`concept_id = 21492239`): get its safe statistics and draw its histogram.

```{r e2}
# STEP 1 — disclosure-safe summary (pooled): n, mean, sd, p05-p95
ds.omop.column.stats("measurement", "value_as_number", concept_id = ____,
                     scope = "pooled", symbol = "omop", conns = conns)$pooled

# STEP 2 — its shape: pull the per-site bins, then draw them
h <- ds.omop.value.histogram("measurement", value_col = "value_as_number",
                             concept_id = ____, symbol = "omop", conns = conns)
# combine h$per_site (drop NA counts), put midpoints on a common grid, then barplot()
```

**Question:** Is systolic blood pressure roughly symmetric, or skewed like creatinine?

::: {.callout-tip collapse="true"}
## Hint
Use `concept_id = 21492239` in both calls. For the picture, reuse the tutorial's `draw_hist()` helper (copy it across), e.g. `draw_hist(21492239, nbins = 9, xlab = "Systolic BP (mmHg)", main = "Systolic BP")`. Its mean is ~115 mmHg.
:::

# Exercise 3 — A new categorical variable

The tutorial counted heart **rhythm**. Explore **marital status** (`concept_id = 40766231`, in the `observation` table): count how often each value occurs.

```{r e3}
# STEP — marital status is a CODED value, so it lives in value_as_concept_id. Count it, pooled.
ds.omop.value.counts("observation", "____", concept_id = 40766231,
                     scope = "pooled", symbol = "omop", conns = conns)$pooled
```

**Question:** What is the most common marital status?

::: {.callout-tip collapse="true"}
## Hint
The column is `"value_as_concept_id"` (categorical measurements/observations store their code there, not in `value_as_number`).
:::

# Exercise 4 — Build your own recipe

Build a wide person-level table `D` with: **sex**, **age** (anchored to 2024), the **mean systolic blood pressure**, a **binary alcohol-abuse flag** (the concept from Exercise 1), and a **count of hospital visits**. Then ratify what the server built and summarise it.

```{r e4}
# STEP 1 — write the recipe: fill the format of each variable and the age anchor year
rec <- omop_recipe(
  variables = list(
    omop_variable(table = "person", column = "gender_concept_id", format = "sex_mf", name = "sex"),
    omop_variable_age(name = "age", year = ____),
    omop_variable(table = "measurement", concept_id = 21492239, format = "____", name = "systolic_bp"),
    omop_variable(table = "condition_occurrence", concept_id = 433753, format = "____", name = "alcohol_abuse"),
    omop_variable(table = "visit_occurrence", format = "count", name = "n_visits")
  ),
  output = omop_output(name = "study", type = "wide"))

# STEP 2 — execute into a server symbol D, then RATIFY it was built as you expected
recipe_execute(rec, out = c(study = "D"), symbol = "omop", conns = conns)
ds.colnames("D", datasources = conns)
ds.dim("D", datasources = conns)

# STEP 3 — summarise the systolic_bp column
ds.summary("D$systolic_bp", datasources = conns)[[1]][["quantiles & mean"]]
```

::: {.callout-tip collapse="true"}
## Hint
`year = 2024`; a *mean* lab value is `format = "mean"`; a yes/no condition flag is `format = "binary"`. `D` should have one row per patient with `person_id, sex, age, systolic_bp, alcohol_abuse, n_visits`.
:::

# Exercise 5 — Filter at extraction

Re-extract, but keep only **male** patients **aged 60+**, as `DSUB`. Confirm the cohort shrank and that everyone is male.

```{r e5}
# STEP — add a filters = list(...) with a sex filter and an age filter
rec_sub <- omop_recipe(
  variables = list(
    omop_variable(table = "person", column = "gender_concept_id", format = "sex_mf", name = "sex"),
    omop_variable_age(name = "age", year = 2024),
    omop_variable(table = "condition_occurrence", concept_id = 433753, format = "binary", name = "alcohol_abuse")
  ),
  filters = list(
    male  = omop_filter_sex("____"),
    older = omop_filter_age(min = ____, year = 2024)
  ),
  output = omop_output(name = "study", type = "wide"))

recipe_execute(rec_sub, out = c(study = "DSUB"), symbol = "omop", conns = conns)
ds.dim("DSUB", datasources = conns)
as.data.frame(ds.table("DSUB$sex", datasources = conns)$output.list[["TABLES.COMBINED_all.sources_counts"]])
```

**Question:** How many patients remain, and is every one of them male?

::: {.callout-tip collapse="true"}
## Hint
`omop_filter_sex("M")` and `omop_filter_age(min = 60, year = 2024)`. Filters can only shrink a population, so the count drops.
:::

# Exercise 6 — A model

Build a table with **sex**, **age** and a **binary hypertension** flag (`concept_id = 320128`), then fit a logistic regression of hypertension on age and sex.

```{r e6}
# STEP 1 — recipe + execute into H
rec_h <- omop_recipe(
  variables = list(
    omop_variable(table = "person", column = "gender_concept_id", format = "sex_mf", name = "sex"),
    omop_variable_age(name = "age", year = 2024),
    omop_variable(table = "condition_occurrence", concept_id = 320128, format = "binary", name = "hypertension")
  ),
  output = omop_output(name = "study", type = "wide"))
recipe_execute(rec_h, out = c(study = "H"), symbol = "omop", conns = conns)

# STEP 2 — fit the logistic model: fill in the two predictors
fit <- ds.glm(formula = "H$hypertension ~ H$____ + H$____",
              family = "binomial", datasources = conns)
fit$coefficients
```

**Question:** Which predictor is significant (p < 0.05)? Is the direction clinically sensible?

::: {.callout-tip collapse="true"}
## Hint
The predictors are `H$age` and `H$sex`. Read the `p-value` column; `age` should be the significant one (odds of hypertension rise with age). On ~100 patients, expect wide confidence intervals.
:::

# Open challenge

Pick **any** condition that interests you (find it with `ds.omop.concept.search`), add it to a recipe as a binary flag, and explore whether it is associated with age or sex. There is no single right answer — the skill is running the workflow correctly and reading the result honestly.

```{r challenge}
# YOUR CODE HERE
```

```{r logout}
DSI::datashield.logout(conns)
```

::: {.callout-note}
Not every hypothesis pans out — a non-significant result is still a real finding. Doing the analysis correctly matters more than getting stars next to your coefficients.
:::
